---
title: "12_Spatial_Transcriptomics"
author: "Aylin Henstridge (Aylinhen@umich.edu)"
date: "2026-06-24"
output: 
  html_document:
    toc: true
    toc_depth: 2
    warnings: FALSE
    number_sections: true
    toc_float: true
    code_download: true
    fig_caption: true
    fig_width: 10
    fig_height: 6
    highlight: 'tango'
    theme: united
knit: (function(input, ...){rmarkdown::render(input, output_dir = "../DOCUMENTS")})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(toc = TRUE)
```

```{css, echo = FALSE}
pre {
  max-height: 500px;
  overflow-y: auto;
}

pre[class]{
  max-height: 500px;
}

.scroll-100{
  max-height: 500px;
  overflow-y: auto;
  background-color: inherit;
}
```

# Load in libraries and data
```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
library(Seurat)
library(SeuratObject)
library(SeuratWrappers)
library(AUCell)
library(monocle3)
library(GSVA)
library(pheatmap)
library(tidyverse)
library(RColorBrewer)
library(DT)
library(dplyr)
`%nin%` <- Negate(`%in%`)
getPalette <- colorRampPalette(brewer.pal(9, 'Set1'))
source("../UTILS/pseudobulk_functions.R")
})
```

# Open Spatial Object (Pei et al, 2025, Nature)
```{r}
PDAC_combined <- readRDS("../OBJECTS/PDAC_combined_cytospace_deconvolution.rds") #read in deconvoluted object
#object was shared with us from published data  from DOI: 10.1038/s41586-025-08927-x
```

## Subset cell types of interest from the deconvoluted object
We want to subset out CAFs and epithelial cells from the object.

```{r, results='hide', warning=FALSE}
PDAC_combined$celltype_cytospace <- ifelse(PDAC_combined$CAFS >= 0.8, "CAF",
                                 ifelse(PDAC_combined$Epicells >= 0.8, "Epi", "Other"))

Idents(PDAC_combined) <- "celltype_cytospace"
levels(PDAC_combined) #We now have new metadata that says we have CAFs and EPi cells
table(PDAC_combined$celltype_cytospace) #and it looks like we have 13376 CAFs and 3196 epithelial cells 

caf_epi_cytospace <- subset(PDAC_combined, idents = c("CAF", "Epi"))
```

```{r, echo=FALSE}
DimPlot(object =PDAC_combined, reduction = "umap", pt.size = 0.05, raster = F, label = T, repel = T, cols = getPalette(length(levels(PDAC_combined)))) 
DimPlot(object =caf_epi_cytospace, reduction = "umap", pt.size = 0.05, raster = F, label = T, repel = T, cols = getPalette(length(levels(PDAC_combined)))) 
```

### SUBSET CAFS

We will update CAF metadata and then map agCAF gene expression in the different sites. 

```{r, results='hide', warning=FALSE}
CAFs <- subset(caf_epi_cytospace, idents = c("CAF"))
DimPlot(object =CAFs, reduction = "umap", pt.size = 0.05, raster = F, label = T, repel = T, cols = getPalette(length(levels(PDAC_combined)))) 

DefaultAssay(CAFs) <- "Spatial" 
CAFs$site[CAFs$site == "PC"] <- "Peritoneal Metastases"
CAFs$site[CAFs$site == "Lung"] <- "Lung Metastases"
CAFs$site[CAFs$site == "Liver"] <- "Liver Metastases"
CAFs$site[CAFs$site == "Panc"] <- "Primary Tumor"


genes <- c("PDGFRA", "NEGR1", "SLIT2", "ROBO1", 
           "SEMA3A", "NRP1", "NFIA",
           "EPHA3", "FAP",
           "ACTA2", "CXCL12")
Idents(CAFs) <- "site"
```

```{r, echo=FALSE}
DotPlot(
  object = CAFs,
  features = genes,
  scale = TRUE
) +
  scale_color_gradientn(
    colours = rev(brewer.pal(11, "RdYlBu"))
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, size = 10, hjust = 1)
  )
```

#### AUCELL ANALYSIS ON CAFs USING KNOWN GENE SIGNATURES
```{r, results='hide', warning=FALSE}
library(readxl)
mycaf_icaf <- read_excel("../GENE_SIGNATURES/MYCAF_ICAF.xlsx")
mycaf <- mycaf_icaf$mycaf
icaf <- mycaf_icaf$icaf
apcaf <- mycaf_icaf$apcaf

mycaf_icaf <- list(
  mycaf = mycaf,
  icaf = icaf,
  apcaf = apcaf
)
expr <- GetAssayData(CAFs,
                     assay = "Spatial",
                     layer = "counts")
geneRanking <- AUCell_buildRankings(expr, nCores = 36, splitByBlocks = TRUE)

auc_score_2 <- AUCell_calcAUC(rankings = geneRanking, geneSets = mycaf_icaf, aucMaxRank = nrow(geneRanking)*0.1)

#make a data frame with ALL of the different types of auc scoring generated with different gene sets
auc_score_2 <- as.data.frame(t(assay(c(auc_score_2))))
auc_score_scaled_2 <- as.data.frame(scale(auc_score_2))
#scaled the data and add it as metadata
CAFs <- AddMetaData(CAFs, metadata = auc_score_scaled_2)

# Create the plot object and display it in the console
FeaturePlot(CAFs, features = c("mycaf", "icaf", "apcaf"), label = FALSE, repel = TRUE, order = TRUE) & 
  scale_colour_gradientn(scale = TRUE, limits = c(-2,2), colours = rev(brewer.pal(n = 11, name = "RdYlBu")))
```

```{r, results='hide', warning=FALSE}

###___________________________________________________
agf_signature <- c("NEGR1", "SLIT2", "ROBO1", "EPHA3","NRP1", "NFIA", 'SEMA3A')
AGF <- list(
  
  agf_signature = agf_signature
)
expr <- GetAssayData(CAFs,
                     assay = "Spatial",
                     layer = "counts")

geneRanking <- AUCell_buildRankings(expr, nCores = 36, splitByBlocks = TRUE)
auc_score <- AUCell_calcAUC(rankings = geneRanking, geneSets = AGF, aucMaxRank = nrow(geneRanking)*0.1)
#make a data frame with ALL of the different types of auc scoring generated with different gene sets
auc_score <- as.data.frame(t(assay(c(auc_score))))
auc_score_scaled <- as.data.frame(scale(auc_score))

#scaled the data and add it as metadata
CAFs <- AddMetaData(CAFs, metadata = auc_score_scaled)
FeaturePlot(CAFs, features = c("mycaf", "icaf", "agf_signature"), pt.size =1, label = FALSE, repel = TRUE, order = TRUE) & 
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlBu")))

# 1. Define what "top expressors" means (e.g., top 10% of cells)
agf_threshold <- quantile(CAFs$agf_signature, 0.75)  # top 25%
# 2. Create a new metadata column
CAFs$agf <- ifelse(CAFs$agf_signature >= agf_threshold, "agf", "caf")
# Check distribution
table(CAFs$agf)
```

# Distance Plots
```{r, results='hide', warning=FALSE}

##Find out which tumor cells are near CAFs in the whole PDAC combined objext
Idents(PDAC_combined) <- "site"
levels(PDAC_combined)
PDAC_combined <- subset(PDAC_combined, idents = c("Panc"))
Idents(PDAC_combined) <- "celltype_cytospace"

#We can apply labels to the 'other' cell types downstream for now will focus on CAF and Epi. 
NR_CAFs <- CAFs@meta.data[, "agf", drop = FALSE]
PDAC_combined$celltype_cytospace_agf <- as.character(PDAC_combined$celltype_cytospace)  # start with broad labels
shared_barcodes <- intersect(rownames(NR_CAFs), colnames(PDAC_combined))
PDAC_combined$celltype_cytospace_agf[shared_barcodes] <- NR_CAFs[shared_barcodes, "agf"]
table(PDAC_combined$celltype_cytospace_agf) #these now have Neural CAFs labeled as agf cells and the pther CAFs as caf

```

```{r, results='hide', warning=FALSE}
NRCAF_Spots <- WhichCells(PDAC_combined, expression = celltype_cytospace_agf == "agf") #these are our neural CAF spots
Epi_Spots <- WhichCells(PDAC_combined, expression = celltype_cytospace_agf == "Epi") #these are tumor cells 

#Need to get coordnates from ALL tissue slices
coords_full <- do.call(rbind, lapply(Images(PDAC_combined), function(img) {
  GetTissueCoordinates(PDAC_combined, image = img)
}))

# Verify
nrow(coords_full)
ncol(PDAC_combined)  # should be close, now we know we have the coordinates for all of our image slices 
NRCAF_Spots <- WhichCells(PDAC_combined, expression = celltype_cytospace_agf == "agf")
Epi_Spots   <- WhichCells(PDAC_combined, expression = celltype_cytospace_agf == "Epi")
sum(!NRCAF_Spots %in% rownames(coords_full))  # should be 0
sum(!Epi_Spots   %in% rownames(coords_full))  # should be 0
coords_full$sample <- gsub(".*-1_", "", rownames(coords_full))

# Check samples
unique(coords_full$sample)
```

```{r, results='hide', warning=FALSE}
#we need to confirm the size of pixels in the visium data coordinates
# Visium spot diameter = 55 µm (fixed by 10x Genomics)
# We use the spot scale factor to calculate pixels per spot diameter,
# then derive microns per pixel for biologically interpretable thresholds.

img <- PDAC_combined@images[[1]]
spot_px <- img@scale.factors$spot * ncol(img@image)  # spot diameter in pixels
microns_per_pixel <- 55 / spot_px                          # µm per pixel

cat(sprintf("Spot diameter:      %.1f pixels = 55 µm\n", spot_px))
cat(sprintf("Scale:              %.3f µm per pixel\n",   microns_per_pixel))
cat(sprintf("                    %.3f pixels per µm\n",  1 / microns_per_pixel))

#Collect nearest-neighbor distances (Epi -> NR-CAF) across all
#slices, then use the 25th/75th percentile as near/far cutoffs.
samples   <- unique(coords_full$sample)
all_dists <- c()

for (s in samples) {
  A_s <- NRCAF_Spots[grepl(paste0("_", s, "$"), NRCAF_Spots)]
  B_s <- Epi_Spots[grepl(paste0("_", s, "$"),   Epi_Spots)]
  if (length(A_s) < 3 || length(B_s) < 3) next
  
  d <- FNN::get.knnx(
    data  = as.matrix(coords_full[A_s, c("imagerow", "imagecol")]),
    query = as.matrix(coords_full[B_s, c("imagerow", "imagecol")]),
    k = 1
  )$nn.dist[, 1]
  
  all_dists <- c(all_dists, d)
}

near_cutoff_px <- quantile(all_dists, 0.25) #closest 25%
far_cutoff_px  <- quantile(all_dists, 0.75) #farthest 25%

near_cutoff_um <- near_cutoff_px * microns_per_pixel
far_cutoff_um  <- far_cutoff_px  * microns_per_pixel

near_cutoff_spots <- near_cutoff_px / spot_px
far_cutoff_spots  <- far_cutoff_px  / spot_px

q <- quantile(all_dists * microns_per_pixel,
              probs = c(0.10, 0.25, 0.50, 0.75, 0.90))
print(round(q, 1))

cat(sprintf("Near cutoff (25th pct): <= %.1f px  /  <= %.1f µm  /  <= %.2f spot diameters\n",
            near_cutoff_px, near_cutoff_um, near_cutoff_spots))
cat(sprintf("Far  cutoff (75th pct): >= %.1f px  /  >= %.1f µm  /  >= %.2f spot diameters\n",
            far_cutoff_px,  far_cutoff_um,  far_cutoff_spots))


Epi_near_list <- list()
Epi_far_list  <- list()
skipped       <- c()

for (s in samples) {
  A_s <- NRCAF_Spots[grepl(paste0("_", s, "$"), NRCAF_Spots)]
  B_s <- Epi_Spots[grepl(paste0("_", s, "$"),   Epi_Spots)]
  
  if (length(A_s) < 3 || length(B_s) < 3) {
    skipped <- c(skipped, s)
    next
  }
  
  dists <- FNN::get.knnx(
    data  = as.matrix(coords_full[A_s, c("imagerow", "imagecol")]),
    query = as.matrix(coords_full[B_s, c("imagerow", "imagecol")]),
    k = 1
  )$nn.dist[, 1]
  
  Epi_near_list[[s]] <- B_s[dists <= near_cutoff_px]
  Epi_far_list[[s]]  <- B_s[dists >= far_cutoff_px]
  
  cat(sprintf("  Sample %-6s | Total Epi: %4d | Near: %4d | Far: %4d\n",
              s, length(B_s),
              length(Epi_near_list[[s]]),
              length(Epi_far_list[[s]])))
}

if (length(skipped) > 0) {
  cat("\nSkipped (< 3 NR-CAF or Epi spots):", paste(skipped, collapse = ", "), "\n")
}

Epi_near_NRCAF <- unlist(Epi_near_list)
Epi_far_NRCAF  <- unlist(Epi_far_list)

cat(sprintf("Epi near NR-CAF: %d spots\n", length(Epi_near_NRCAF)))
cat(sprintf("Epi far NR-CAF:  %d spots\n", length(Epi_far_NRCAF)))
cat(sprintf("Total Epi:       %d spots\n", length(Epi_Spots)))
cat(sprintf("Excluded (middle 50%%): %d spots\n",
            length(Epi_Spots) - length(Epi_near_NRCAF) - length(Epi_far_NRCAF)))

# Add labels to metadata
PDAC_combined$Epi_NRCAF_group <- "Intermediate"
PDAC_combined$Epi_NRCAF_group[Epi_near_NRCAF] <- "Near"
PDAC_combined$Epi_NRCAF_group[Epi_far_NRCAF]  <- "Far"

print(table(PDAC_combined$Epi_NRCAF_group, useNA = "ifany"))
```


```{r, results='hide', warning=FALSE}

#Receptors that interact with NR-CAFs
NRP1 <- c("NRP1", "VEGFA",
          "SEMA3B", "SEMA3A", "SEMA3C", "PLXND1")

nc_WNT <- c("WNT5A", "WNT11",
            "FZD2", "FZD5", "FZD7",
            "ROR1", "ROR2", "RYK")

c_WNT <- c("WNT2", "WNT2B","WNT7A", "WNT3A", 
           "WNT7B", "WNT9A", "WNT10A",
           "FZD2", "FZD4", "FZD5", "FZD7", "FZD1", "FZD8", "LRP5", "LRP6")

hedgehog <- c("SHH", "DHH", "IHH", "PTCH1", "PTCH2")

NEGR1 <- c("FGFR2", "ADAM10", "NEGR1")


library(RColorBrewer)
Idents(PDAC_combined) <- "site"
levels(PDAC_combined)

PDAC_combined <- subset(PDAC_combined, idents = c("Panc"))
levels(PDAC_combined)
Idents(PDAC_combined) <- "Epi_NRCAF_group"
levels(PDAC_combined)
```


# Line plots 
```{r}
# Get coordinates
coords_Epi   <- coords_full[Epi_Spots,   c("imagerow", "imagecol")]
coords_NRCAF <- coords_full[NRCAF_Spots, c("imagerow", "imagecol")]

# For each Epi spot, find minimum distance to any NR-CAF
library(FNN)
dist_to_NRCAF <- FNN::get.knnx(
  data  = as.matrix(coords_NRCAF),
  query = as.matrix(coords_Epi),
  k = 1
)$nn.dist[, 1]

# Convert to microns
dist_to_NRCAF_um <- dist_to_NRCAF * microns_per_pixel

# Store in a data frame
epi_dist_df <- data.frame(
  barcode  = Epi_Spots,
  dist_um  = dist_to_NRCAF_um,
  dist_sqrt = sqrt(dist_to_NRCAF_um),  # sqrt transform like the paper
  row.names = Epi_Spots
)

sig_list <- list( NRP1,
                  nc_WNT,
                  NEGR1
)

PDAC_combined <- AddModuleScore(
  PDAC_combined,
  features = sig_list,
  name     = "Sig"  # will create Sig1, Sig2, Sig3, Sig4
)

# STEP 2: Pull scores for Epi spots into the distance dataframe
epi_dist_df$NRP1_Signaling <- PDAC_combined$Sig1[Epi_Spots]
epi_dist_df$non_canonical_Wnt_Signaling     <- PDAC_combined$Sig2[Epi_Spots]
epi_dist_df$NEGR1_Signaling     <- PDAC_combined$Sig3[Epi_Spots]

# STEP 3: Reshape to long format
library(tidyr)

epi_long <- epi_dist_df %>%
  pivot_longer(
    cols      = c("NRP1_Signaling", "non_canonical_Wnt_Signaling", "NEGR1_Signaling"),
    names_to  = "Signature",
    values_to = "Score"
  ) %>%
  mutate(Signature = factor(Signature, 
                            levels = c("NRP1_Signaling", "non_canonical_Wnt_Signaling", 
                                    "NEGR1_Signaling" )))

# STEP 4: Plot all 4 signatures
ggplot(epi_long, aes(x = dist_sqrt, y = Score, color = Signature)) +
  coord_cartesian(xlim = c(0, 4.5)) +
  geom_smooth(
    method    = "loess",
    span      = 0.3,
    se        = TRUE,
    alpha     = 0.15,
    linewidth = 1.2
  ) +
  scale_color_manual(values = c(
    "NRP1_Signaling" = "#4A90C4",   # blue
    "non_canonical_Wnt_Signaling"     = "#8B3A14",   # brown
    "Hedgehog_Signaling"     = "#2E8B57",   # green
    "NEGR1_Signaling"     = "#9B59B6" 
  )) +
  scale_fill_manual(values = c(
    "NRP1_Signaling" = "#4A90C4",   # blue
    "non_canonical_Wnt_Signaling"     = "#8B3A14",   # brown
    "Hedgehog_Signaling"     = "#2E8B57",   # green
    "NEGR1_Signaling"     = "#9B59B6" 
  )) +
  labs(
    title    = "Epithelial Signature Scores vs. Distance from NR-CAFs",
    subtitle = "Loess smoothing; shaded area = 95% CI",
    x        = "Square Root (Radial Distance from NR-CAF, µm)",
    y        = "Signature Score",
    color    = "Signature"
  ) +
  theme_classic() +
  theme(
    plot.title      = element_text(size = 12, face = "bold"),
    plot.subtitle   = element_text(size = 10, color = "grey40"),
    axis.title      = element_text(size = 11),
    legend.position = "right",
    legend.title    = element_text(size = 10, face = "bold"),
    legend.text     = element_text(size = 9)
  )

```

```{r}
sig_list <- list( "NRP1",
                  "VEGFA",
                  "SEMA3A",
                  "SEMA3B",
                  "SEMA3C"
)

PDAC_combined <- AddModuleScore(
  PDAC_combined,
  features = sig_list,
  name     = "NRP1_RECEPTORS"  # will create Sig1, Sig2, Sig3, Sig4
)

# STEP 2: Pull scores for Epi spots into the distance dataframe
epi_dist_df$NRP1 <- PDAC_combined$NRP1_RECEPTORS1[Epi_Spots]
epi_dist_df$VEGFA     <- PDAC_combined$NRP1_RECEPTORS2[Epi_Spots]
epi_dist_df$SEMA3A <- PDAC_combined$NRP1_RECEPTORS3[Epi_Spots]
epi_dist_df$SEMA3B     <- PDAC_combined$NRP1_RECEPTORS4[Epi_Spots]
epi_dist_df$SEMA3C     <- PDAC_combined$NRP1_RECEPTORS5[Epi_Spots]


# STEP 3: Reshape to long format
library(tidyr)

epi_long <- epi_dist_df %>%
  pivot_longer(
    cols      = c("NRP1",
                  "VEGFA",
                  "SEMA3A",
                  "SEMA3B",
                  "SEMA3C"),
    names_to  = "Signature",
    values_to = "Score"
  ) %>%
  mutate(Signature = factor(Signature, 
                            levels = c("NRP1",
                                       "VEGFA",
                                       "SEMA3A",
                                       "SEMA3B",
                                       "SEMA3C")))

# STEP 4: Plot all 4 signatures
ggplot(epi_long, aes(x = dist_sqrt, y = Score, color = Signature)) +
  geom_smooth(
    method    = "loess",
    span      = 0.3,
    se        = TRUE,
    alpha     = 0.15,
    linewidth = 1.2
  ) +
  coord_cartesian(xlim = c(0, 4.5)) +
  scale_color_manual(values = c(
    "NRP1" = "#4A90C4",   # blue
    "VEGFA"     = "#8B3A14",   # brown
    "SEMA3A"     = "#9B59B6" ,
    "SEMA3B" = "pink",
    "SEMA3C" = "red"
  )) +
  scale_fill_manual(values = c(
    "NRP1" = "#4A90C4",   # blue
    "VEGFA"     = "#8B3A14",   # brown
    "SEMA3A"     = "#9B59B6" ,
    "SEMA3B" = "pink",
    "SEMA3C" = "red"
  )) +
  labs(
    title    = "Epithelial NRP1 Receptor Gene Expression Scores vs. Distance from NR-CAFs",
    subtitle = "Loess smoothing; shaded area = 95% CI",
    x        = "Square Root (Radial Distance from NR-CAF, µm)",
    y        = "Signature Score",
    color    = "Signature"
  ) +
  theme_classic() +
  theme(
    plot.title      = element_text(size = 12, face = "bold"),
    plot.subtitle   = element_text(size = 10, color = "grey40"),
    axis.title      = element_text(size = 11),
    legend.position = "right",
    legend.title    = element_text(size = 10, face = "bold"),
    legend.text     = element_text(size = 9)
  )


```

