---
title: "01_Generating_Merged_Object"
author: "Aylin Henstridge (Aylinhen@umich.edu)"
date: "2026-02-11"
output: 
  html_document:
    toc: true
    toc_depth: 2
    warnings: FALSE
    number_sections: true
    toc_float: true
    code_download: true
    fig_caption: true
    fig_width: 10
    fig_height: 6
    highlight: 'tango'
    theme: united
knit: (function(input, ...){rmarkdown::render(input, output_dir = "../DOCUMENTS")})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(toc = TRUE)
```

```{css, echo = FALSE}
pre {
  max-height: 500px;
  overflow-y: auto;
}

pre[class]{
  max-height: 500px;
}

.scroll-100{
  max-height: 500px;
  overflow-y: auto;
  background-color: inherit;
}
```

# Generating Merged Longitduinal Object with Corrected Counts
## Loading Libraries
```{r, message=FALSE, warning=FALSE}
###########################################################################
#              Generating Merged Longitudinal Object                      #
############################################################################
library(SeuratObject)
library(parallel)
library(tidyverse)
library(ggplot2)
library(magrittr)
source("../UTILS/correct_reads_function.r")
source("../UTILS/pseudobulk_functions.R")
library(Seurat)
library(dplyr)
library(data.table)
library(Matrix)
library(devtools)
library(RcppArmadillo)
library(Rcpp)
library(scales)
library(pheatmap)
library(gplots)
library(cowplot)
library(RColorBrewer)
library(viridis)
`%nin%` <- Negate(`%in%`)
getPalette <- colorRampPalette(brewer.pal(9, 'Set1'))
```

## Read in data
 You will use the data that was generated in the previous script (# 00) --> temp.rds is a lsit of all of your obejcts that we will
 split, normalzie and merge as shown below. We will generate the SCT and Integration assays as we do this. 
```{r, eval=FALSE} 

tumor_obj <- readRDS("../OBJECTS/temp.rds") #this is the list we generated in script 00

#Samples integration
features <- SelectIntegrationFeatures(object.list = tumor_obj, nfeatures = 3000)
tumor_obj <- PrepSCTIntegration(object.list = tumor_obj, anchor.features = features)
anchors <- FindIntegrationAnchors(object.list = tumor_obj, normalization.method = "SCT",
                                  anchor.features = features)
tumor_obj <- IntegrateData(anchorset = anchors, normalization.method = "SCT") #this step takes quite a few hours, best to run over night 

saveRDS(tumor_obj, "../OBJECTS/Merged_Longitudinal_Object_AH1.rds")
```

# Annotating Merged Object
Merged Longitudinal Object AH1 is the merged object that we will now annotate, subset, and analyze.

## Naming Clusters
```{r, message=FALSE, warning=FALSE}
##########################################################################
#                                  Naming Clusters                      #
############################################################################

longitudinal <- readRDS("../OBJECTS/Merged_Longitudinal_Object_AH1.rds")

DefaultAssay(longitudinal) <- "integrated"
# dimension reduction and clustering of integrated object
set.seed(42)
longitudinal <- RunPCA(longitudinal, verbose = FALSE)
longitudinal <- RunUMAP(longitudinal, reduction = "pca", dims = 1:30, verbose = FALSE)
longitudinal <- FindNeighbors(longitudinal, reduction = "pca", dims = 1:30)
longitudinal <- FindClusters(longitudinal, resolution = 1.8)
longitudinal[["seurat_clusters"]] <- Idents(object = longitudinal)

#Look at what clusters look like at this step
DimPlot(object =longitudinal, reduction = "umap", pt.size = 0.1, label = T, repel = T)

Idents(object = longitudinal) <- 'seurat_clusters'
longitudinal <- PrepSCTFindMarkers(longitudinal, assay = "SCT", verbose = TRUE)
markers <- FindAllMarkers(longitudinal, assay = 'SCT', min.pct = 0.25, only.pos = T)
write.csv(markers, file = "../DATA/longitudinal.markers.csv")
```

### Feature plots used to name clusters.
```{r, message=FALSE, warning=FALSE}
#Name the clusters, we have 47 clusters and will use common gene markers to identify them. 
DefaultAssay(longitudinal) <- "SCT"
FeaturePlot(object = longitudinal, features = c("EPCAM","KRT19", "MUC1", "KRT18"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order =T) #Epithelial
FeaturePlot(object = longitudinal, features = c('PRSS1', 'AMY2A'), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order =T) #acinar
FeaturePlot(object = longitudinal, features = c("CDH5", "VWF"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01,  order =T) #Endothelial. 
FeaturePlot(object = longitudinal, features = c("PDGFRB", "PDGFRA", "PDPN", "ACTA2", "RGS5","PECAM1", "COL1A1", "IGFBP7"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #Fibroblasts and pericyte
FeaturePlot(object = longitudinal, features = c("PTPRC", "CD14", "ITGAM", "MARCO", "APOE", "C1QA", "S100A8", "FCGR3A", "FCGR3B", "SPP1", "CD68", "HLA-DRA", 'G0S2'), cols = c("grey", "red"), reduction = "umap", pt.size = .05, order = T) #Myeloid
FeaturePlot(object = longitudinal, features = c("CD3E", "CD4", "FOXP3", "CD8A", "GZMB", "PRF1", "NKG7", "NCAM1", "MKI67"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #T cells
FeaturePlot(object = longitudinal, features = c("CD19", "CD79A", "IGJ", "IGLL5", "CD27", "CD138", 'MS4A1', 'JCHAIN'), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #B cells
FeaturePlot(object = longitudinal, features = c("HBA2", "HBA1"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #RBCs
FeaturePlot(object = longitudinal, features = c("CLU", "TPSAB1", "CPA3"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #Mast cells
FeaturePlot(object = longitudinal, features = c("MKI67"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #Cycling
FeaturePlot(object = longitudinal, features = c('IRF7', 'GZMB', 'CLEC4C'), cols = c("grey", "red"), reduction = "umap", pt.size = .01, order = T) #pDCs
FeaturePlot(object = longitudinal, features = c('PPBP', 'PF4', 'CLU'), cols = c("grey", "red"), reduction = "umap", pt.size = .01, order = T) #platelet
FeaturePlot(object = longitudinal, features = c('KRT20', 'CEACAM1', 'CA2', 'INS'), cols = c("grey", "red"), reduction = "umap", pt.size = .01, order = T) #platelet
```

```{r, message=FALSE, warning=FALSE}
Idents(object = longitudinal) <- 'seurat_clusters'
levels(longitudinal)
current.cluster.ids <- c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25,26, 27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,43,44,45,46, 47)
new.cluster.ids <- c('Granulocyte', 'Granulocyte', 'CD8T', 'CD8T', 'Granulocyte', 'Granulocyte', 
                     'Macrophage', 'Granulocyte', 'Fibroblast', 'Macrophage', 'NK', 
                     'CD4T', 'Acinar', 'CD4T', 'Macrophage', 'RBC', 
                     'Platelet', 'CD4T', 'Epi', 'B Cell', 'Endothelial', 
                     'Granulocyte', 'Granulocyte', 'Fibroblast', 'Acinar', 'Pericyte', 
                     'Junk', 'Fibroblast', 'RBC', 'CD8T', 'Epi', 
                     'Macrophage','NK', 'TReg', 'Junk', 'Macrophage', 
                     'CD4T', 'Macrophage', 'RBC', 'Cycling', 'Mast', 
                     'Junk', 'Plasma', 'CD8T', 'Cycling', 'Platelet','Junk','Plasma')
names(x = new.cluster.ids) <- levels(x = longitudinal)
longitudinal<- RenameIdents(object = longitudinal, new.cluster.ids)

DimPlot(object =longitudinal, reduction = "umap", pt.size = 0.1, label = T, repel = T)
longitudinal[["Level1Labels"]] <- Idents(object = longitudinal)
Idents(object = longitudinal) <- 'Level1Labels'
levels(longitudinal)

longitudinal_fil <- subset(longitudinal, idents = c('Junk', 'RBC'), invert = TRUE)
Idents(object = longitudinal_fil) <- 'Level1Labels'
DimPlot(object =longitudinal_fil, reduction = "umap", pt.size = 0.1, label = F, repel = T, cols = getPalette(length(levels(longitudinal))), split.by= "timepoint")
```

```{r, message=FALSE, warning=FALSE}
DimPlot(object =longitudinal_fil, reduction = "umap", pt.size = 0.01, label = F, repel = T, split.by= "timepoint",  cols = c("CD8T" = "aquamarine2", 
                                                                                                                              "CD4T" = "coral1",
                                                                                                                              "TReg" = "sienna3",
                                                                                                                              "Cycling" = "violetred",
                                                                                                                              "Pericyte" = "lavender",
                                                                                                                              "Macrophage" = "peachpuff", 
                                                                                                                              "Platelet" = "yellow", 
                                                                                                                              "Mast" = "chartreuse1", 
                                                                                                                              "NK" = "seagreen4", 
                                                                                                                              "B Cell" = "lightskyblue1", 
                                                                                                                              "Plasma" = "dodgerblue3", 
                                                                                                                              "Fibroblast" = "goldenrod", 
                                                                                                                              "Endothelial" = "magenta", 
                                                                                                                              'Epi' = "mediumorchid4",
                                                                                                                              'Granulocyte' = 'wheat3',
                                                                                                                              "Acinar" = 'thistle'))
```

```{r, message=FALSE, warning=FALSE}

Idents(longitudinal_fil) <- "Level1Labels"

Idents(longitudinal_fil) <- factor(Idents(longitudinal_fil), levels = rev(c('Epi', 'Acinar', 'Fibroblast', 'Pericyte', 'Endothelial', 'Macrophage', 'Granulocyte',  'Mast', 'NK', 'CD8T',  'CD4T', 'TReg', 
                                                                            'B Cell', 'Plasma','Platelet', 'Cycling')))


DotPlot(object = longitudinal_fil, features = c("EPCAM",  'MUC1', "KRT19", 'AMY2A', 'PRSS1', 'IGFBP7', 'COL6A2', 'PDGFRA', 'RGS5', 'VWF', 'CDH5', "PTPRC","HLA-DRA", "CD14",  
                                                'FCGR3B', 'G0S2', "TPSAB1", "CPA3", "NKG7","NCAM1", "CD3E","CD8A",  "CD4", 'FOXP3',  "CD79A", "MS4A1",'JCHAIN',  "CLU", 'PF4', 'MKI67'), cols = c("white", "magenta4")) +
  theme(legend.position = "right",axis.text.x = element_text(angle = 45, size = 10,  hjust = 1))
```

## Rename sample_id to patient id and timepoint 
```{r, message=FALSE, warning=FALSE}
Idents(longitudinal_fil) <- "sample_id"
levels(longitudinal_fil)
longitudinal_fil$sample_id[longitudinal_fil$sample_id == "3079-EC-2"] <- "1402_T1"
longitudinal_fil$sample_id[longitudinal_fil$sample_id == "6894-EC-1"] <- "1402_T2"
longitudinal_fil$sample_id[longitudinal_fil$sample_id == "4358-EC-1"] <- "1471_T1"
longitudinal_fil$sample_id[longitudinal_fil$sample_id == "5392-EC-1"] <- "1471_T2"
longitudinal_fil$sample_id[longitudinal_fil$sample_id == "4425-EC-1"] <- "1475_T1"
longitudinal_fil$sample_id[longitudinal_fil$sample_id == "6873-EC-1"] <- "1475_T2"
longitudinal_fil$sample_id[longitudinal_fil$sample_id == "5161-EC-1"] <- "1529_T1"
longitudinal_fil$sample_id[longitudinal_fil$sample_id == "6216-EC-1"] <- "1529_T2"
longitudinal_fil$sample_id[longitudinal_fil$sample_id == "5570-EC-1"] <- "1550_T1"
longitudinal_fil$sample_id[longitudinal_fil$sample_id == "6846-EC-1"] <- "1550_T2"
levels(longitudinal_fil)
```

This Annotated object is what you will focus on downstream. 
```{r, message=FALSE, warning=FALSE, eval=FALSE}
saveRDS(longitudinal_fil, "../OBJECTS/Annotated_Longitudinal_Object_AH.rds") 
```

## Cell frequency histograms
```{r, message=FALSE, warning=FALSE}
#Let's make some histograms

longitudinal_fil$Level1Labels <- droplevels(longitudinal_fil@meta.data$Level1Labels)
#Now let's do a histogram by patient and time point
levels(longitudinal)
Idents(longitudinal) <- factor(Idents(longitudinal), levels = rev(c('Epi', 'Acinar', 'Fibroblast', 'Pericyte', 'Endothelial', 'Macrophage', 'Granulocyte',  'Mast', 'NK', 'CD8T',  'CD4T', 'TReg', 
                                                                    'B Cell', 'Plasma','Platelet', 'Cycling')))
#splitting samples
samples.list <- unique(longitudinal_fil$sample_id)
clusters <- lapply(samples.list, function(x){
  subset <- subset(longitudinal_fil, subset = sample_id == x)
  dist <- data.frame(table(subset$Level1Labels))
  
  return(dist)
})

names(clusters) <- samples.list

#calculate relative freq (fractions) of each cell type
clusters_percent <- lapply(clusters, FUN = function(x){
  summ <- sum(x$Freq)
  x$Freq <- (x$Freq/summ)
  return(x)
})
levels(longitudinal_fil)
#making things ggplot-friendly!
clusters_dist <- reshape2::melt(clusters, id.var = "Var1")
colnames(clusters_dist) <- c("Level1Labels","variable","value","sample_id")   
clusters_percent_dist <- reshape2::melt(clusters_percent, id.var = "Var1")
colnames(clusters_percent_dist) <- c("Level1Labels","variable","value","sample_id")

ggplot(clusters_percent_dist, aes(fill=Level1Labels, y = value, x = sample_id)) + 
  scale_x_discrete(limits = c('1402_T1', '1402_T2', '1550_T1', '1550_T2', '1471_T1', '1471_T2', '1475_T1', '1475_T2', '1529_T1', '1529_T2')) +
  geom_bar(position = "stack", stat = "identity") + 
  theme_bw() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, size = 8, face = "bold", hjust = 1)) +
  ylab("count") + xlab("patient_id") + ggtitle("Relative Cell Types Abundance") +
  scale_fill_manual(values = c("wheat3", "peachpuff","aquamarine2", 'seagreen4','coral1', 'goldenrod', 'mediumorchid4', 'sienna3', 'lightskyblue1', 'yellow', 'magenta', 'lavender', 'thistle', 'violetred', 'dodgerblue3', 'chartreuse1'))
```

# Subsetting the Objecyt
```{r, message=FALSE, warning=FALSE, eval=FALSE}
###########################################################################
#              Subsetting Longitudinal Object                      #
############################################################################
## Based on the 03_SubClustering.Rmd notebook findings, here I will implement the third approach for subsetting 
## a cell-type and re-clustering. After subsetting, we will split the object to the samples, re-run SCTransform, 
## extract variable features and re-run PCA and other steps. No need to re-integrate
# Loading libraries and data ----------------------------------------------
integrated_data <- readRDS("../OBJECTS/Annotated_Longitudinal_Object_AH.rds") #rename longitudinal_fil object integrated data here

subsets <- list('Myeloid' = c("Granulocyte","Macrophage"),
                'Epi' = c("Epi", 'Acinar'),
                'T_Cells' = c('CD8T','NK','CD4T','TReg'),
                'Fibro' = c('Fibroblast','Endothelial','Pericyte'))

for (type in names(subsets)){
  
  message("Subsetting ", type)
  
  subset <- subset(integrated_data, subset = Level1Labels %in% subsets[[type]])

  #We only use corrected counts assy here since thats what we will focus on downstream. 
  
  DefaultAssay(subset) <- "integrated"
  subset <- RunPCA(subset, features = features, verbose = FALSE)
  subset <- RunUMAP(subset, reduction = "pca", dims = 1:25, verbose = FALSE)
  subset <- FindNeighbors(subset, reduction = "pca", dims = 1:25, verbose = FALSE)
  subset <- FindClusters(subset, resolution = 1, verbose = FALSE)

  message('Saving')
  saveRDS(subset, paste0("../OBJECTS/",type,"_subset.rds"))
}
```

# Pseudobulk of All Cells#
```{r, message=FALSE, warning=FALSE, eval=FALSE}
longitudinal <- readRDS("../OBJECTS/Annotated_Longitudinal_Object_AH.rds")
```

```{r, message=FALSE, warning=FALSE}
all_cells_DGE <- pseudobulk_dge(seurat_object = longitudinal, assay = "CorrectedCounts",
                                aggregate_by = "sample_id", comparison = "timepoint")
all_pca <- pca_plot(all_cells_DGE, groups = "timepoint") + ggtitle("All cells")
all_pca
```
