---
title: "03_Merging_PT1475_Recurrence_and_Cell_Frequencies.rmd"
author: "Aylin Henstridge (Aylinhen@umich.edu)"
date: "2025-12-09"
output: 
  html_document:
    toc: true
    toc_depth: 2
    warnings: FALSE
    number_sections: true
    toc_float: true
    code_download: true
    fig_caption: true
    fig_width: 10
    fig_height: 6
    highlight: 'tango'
    theme: united
knit: (function(input, ...){rmarkdown::render(input, output_dir = "../DOCUMENTS")})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(toc = TRUE)
```

```{css, echo = FALSE}
pre {
  max-height: 500px;
  overflow-y: auto;
}

pre[class]{
  max-height: 500px;
}

.scroll-100{
  max-height: 500px;
  overflow-y: auto;
  background-color: inherit;
}
```

# Merging all 3 time points from patient 1475

## Load Libraries and data 
```{r, warning=FALSE, message=FALSE}


#load libraries 
suppressPackageStartupMessages({
library(Seurat)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(ComplexHeatmap)
})

#Read in object that contains data from the time_point3 recurrence 
T3 <- readRDS("../OBJECTS/Patient_1475/T31475.rds")
DefaultAssay(T3) <- "RNA"
T3[["percent.mt"]] <- PercentageFeatureSet(T3, pattern = "^MT-")
T3 <- subset(T3, subset = nFeature_RNA > 200 & percent.mt < 15)
T3  <- NormalizeData(T3, normalization.method = "LogNormalize", scale.factor = 10000)
T3 <- FindVariableFeatures(T3, selection.method = "vst", nfeatures = 2000, assay = "RNA")
T3 <- ScaleData(T3, features = rownames(T3), verbose = FALSE)
T3 <- RunPCA(T3, verbose = FALSE)
ElbowPlot(T3, 50)
PCNum = 25
T3 <- FindNeighbors(T3, reduction = "pca", dims = 1:PCNum)
T3 <- FindClusters(T3, resolution = 1.8)
T3 <- RunUMAP(object = T3,dims = 1:PCNum)
DimPlot(object = T3, reduction = "umap", label = TRUE, pt.size = 0.03)
T3[["UMAP_Clusters"]] <- Idents(object = T3)      
T3$sample_id <- "1475 T3"
T3$time_point <- "3"
Idents(T3) <- 'time_point'
levels(T3)
Idents(T3) <- "sample_id"
levels(T3)
Idents(T3) <- "UMAP_Clusters"
```

## Find Cluster Names for T3

```{r, warning=FALSE, message=FALSE}
T3.markers <- FindAllMarkers(T3, only.pos = TRUE, min.pct = 0.25,logfc.threshold = 0.25) #always use 0.25 for percent and fold change threshold
T3.markers %>% group_by(cluster) %>% top_n(n=2, wt=avg_log2FC) #ranks biggest to smallest
write.csv(T3.markers, file = "../DATA_FILES/T3_CLUSTERS_GENES.csv") #this will create CSV file of clusters and genes        
```

## Plot Gene Expression to Validate
```{r}
FeaturePlot(T3, features = c("CD4", "CD8A", "FN1", "FAP", "HBB","EPCAM", "NRCAM", "GZMK"))
```

## Name Clusters Based on Published Data

```{r, warning=FALSE, message=FALSE}
#TO NAME CLUSTERS:
Idents(object = T3) <- 'UMAP_Clusters'
levels(T3)
current.cluster.ids <- c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14)
new.cluster.ids <- c('tcell','Epi','RBC','Fibroblast','Fibroblast','Fibroblast','Endothelial','Macrophage','tcell','Junk','Nerves','T-Reg','Granulocyte','Epi','Nerves')
names(x = new.cluster.ids) <- levels(x = T3)
T3 <- RenameIdents(object = T3, new.cluster.ids)

#Put these new cluster labels as metadata
T3[["Level1Labels"]] <- Idents(object = T3)
Idents(T3) <- "Level1Labels"
levels(T3)
#re-plot cluster with new cluster names so you can check that it worked:
DimPlot(object = T3, reduction = "umap", label = TRUE, pt.size = 0.01)
#saveRDS(T3, "../OBJECTS/Patient_1475/1475_T3.rds")
```

## Now read in Timepoints 1 & 2 to merge with 3
```{r, warning=FALSE, message=FALSE}
LONG <- readRDS("../OBJECTS/eileen_annotated_longitudinal.rds")
Idents(LONG) <- "sample_id"
levels(LONG)
Idents(LONG) <- "sample_id"
levels(LONG)
new.cluster.ids <- c("1402T1", "1402T2","1471T1","1471T2","1475T1","1475T2","1529T1","1529T2","1550T1","1550T2")
names(X=new.cluster.ids) <- levels(x=LONG)
LONG<- RenameIdents(object = LONG, new.cluster.ids)
LONG[["sample_id"]] <- Idents(object = LONG)
levels(LONG)
PT1475 <- subset(LONG, idents = c("1475T1", "1475T2"))
levels(PT1475)
#saveRDS(PT1475, "../OBJECTS/Patient_1475/PT1475_T1_T2.rds")
```
## SUBSET out the time point 1 and timepoint 2 data from patient 1475
```{r, warning=FALSE, message=FALSE}
DefaultAssay(PT1475) <- "RNA"
PT1475[["percent.mt"]] <- PercentageFeatureSet(PT1475, pattern = "^MT-")
PT1475 <- subset(PT1475, subset = nFeature_RNA > 200 & percent.mt < 15)
PT1475  <- NormalizeData(PT1475, normalization.method = "LogNormalize", scale.factor = 10000)
PT1475 <- FindVariableFeatures(PT1475, selection.method = "vst", nfeatures = 2000, assay = "RNA")
PT1475 <- ScaleData(PT1475, features = rownames(PT1475), verbose = FALSE)
PT1475 <- RunPCA(PT1475, verbose = FALSE)
ElbowPlot(PT1475, 50)
PCNum = 25
PT1475 <- FindNeighbors(PT1475, reduction = "pca", dims = 1:PCNum)
PT1475 <- FindClusters(PT1475, resolution = 1.8)
PT1475 <- RunUMAP(object = PT1475, dims = 1:PCNum)
Idents(PT1475) <- "Level1Labels"
DimPlot(object = PT1475, reduction = "umap", label = TRUE, pt.size = 0.03)
levels(PT1475)
```

```{r, warning=FALSE, message=FALSE}
Idents(T3) <- "Level1Labels"
Idents(PT1475) <- "Level1Labels"

PT1475_FULL <- merge(PT1475, T3, merge.data = TRUE)
Idents(PT1475_FULL) <- "Level1Labels"
levels(PT1475_FULL)

Idents(PT1475_FULL)<- "time_point"
table(Idents(PT1475_FULL))

#saveRDS(PT1475_FULL, "../OBJECTS/Patient_1475/1475_T1_T2_T3.rds")
```

## Now check integration 
```{r, warning=FALSE, message=FALSE}
#PT1475_FULL <- readRDS("../OBJECTS/Patient_1475/1475_T1_T2_T3.rds")

DefaultAssay(PT1475_FULL)<- "RNA"
PT1475_FULL <- NormalizeData(PT1475_FULL)
PT1475_FULL <- FindVariableFeatures(PT1475_FULL, selection.method = "vst", nfeatures = 2000)
PT1475_FULL <- ScaleData(PT1475_FULL)
PT1475_FULL <- RunPCA(PT1475_FULL)
ElbowPlot(PT1475_FULL, 50)
PT1475_FULL <- FindNeighbors(PT1475_FULL, reduction = "pca", dims = 1:25)
PT1475_FULL <- FindClusters(PT1475_FULL, resolution = 1.2, cluster.name = "unintegrated_clusters")

PT1475_FULL <- RunUMAP(PT1475_FULL, dims = 1:25, reduction = "pca", reduction.name = "umap.unintegrated")
DefaultAssay(PT1475_FULL) <- "RNA"
PT1475_FULL <- IntegrateLayers(
  object = PT1475_FULL, method = CCAIntegration,
  orig.reduction = "pca", new.reduction = "integrated.cca",
  verbose = FALSE
)

PT1475_FULL <- FindNeighbors(PT1475_FULL, reduction = "integrated.cca", dims = 1:25)
PT1475_FULL <- FindClusters(PT1475_FULL, resolution = 1.2, cluster.name = "cca_clusters")
PT1475_FULL <- RunUMAP(PT1475_FULL, reduction = "integrated.cca", dims = 1:25, reduction.name = "umap.cca")

Idents(PT1475_FULL) <- "Level1Labels"
levels(PT1475_FULL)
DimPlot(PT1475_FULL,  reduction = "umap.cca",  group.by = "Level1Labels",  combine = FALSE, label.size = 1)
DimPlot(PT1475_FULL,  reduction = "umap.unintegrated",  group.by = c("Level1Labels"),  combine = FALSE, label.size =1)
```

## Pass the labels
```{r, warning=FALSE, message=FALSE}
DefaultAssay(PT1475_FULL) <-'RNA'
DefaultAssay(PT1475) <- 'RNA'
DefaultAssay(T3) <- 'RNA'

MERGED <- rbind(T3@meta.data %>% select(Level1Labels), PT1475@meta.data %>% select(Level1Labels))
MERGED_ALIGNED <- MERGED[rownames(PT1475_FULL@meta.data), ]
PT1475_FULL@meta.data$Level1Labels <- MERGED_ALIGNED

Idents(PT1475_FULL) <- 'Level1Labels'
DimPlot(object =PT1475_FULL, reduction = "umap.cca", pt.size = 0.1, label = T, repel = T, group.by = "Level1Labels", raster = FALSE) +
DimPlot(object =PT1475_FULL, reduction = "umap.unintegrated", pt.size = 0.1, label = T, repel = T, group.by = "Level1Labels", raster = FALSE) 


#Integrated looks cleaner move forward with this version
#USE umap.cca reduction
```

```{r, warning=FALSE, message=FALSE}
DefaultAssay(PT1475_FULL) <- "RNA"
Idents(PT1475_FULL) <- "Level1Labels"

##recluster and clean up labels
merged_clean  <- NormalizeData(PT1475_FULL, normalization.method = "LogNormalize", scale.factor = 10000)
merged_clean <- FindVariableFeatures(merged_clean, selection.method = "vst", nfeatures = 2000, assay = "RNA")
merged_clean <- ScaleData(merged_clean, features = rownames(merged_clean), verbose = FALSE)
merged_clean <- RunPCA(merged_clean, verbose = FALSE)
ElbowPlot(merged_clean, 50)
PCNum = 30
merged_clean <- FindNeighbors(merged_clean, reduction = "pca", dims = 1:PCNum)
merged_clean <- FindClusters(merged_clean, resolution = 1.8)
merged_clean <- RunUMAP(object = merged_clean,dims = 1:PCNum)
DimPlot(object = merged_clean, reduction = "umap", label = TRUE, pt.size = 0.03)
merged_clean[["UMAP_Clusters_new"]] <- Idents(object = merged_clean)      
Idents(merged_clean) <- "UMAP_Clusters_new"
merged_clean <- JoinLayers(merged_clean)
merged_clean.markers <- FindAllMarkers(merged_clean, only.pos = TRUE, min.pct = 0.25,logfc.threshold = 0.25) #always use 0.25 for percent and fold change threshold
merged_clean.markers %>% group_by(cluster) %>% top_n(n=2, wt=avg_log2FC) #ranks biggest to smallest
write.csv(merged_clean.markers, file = "../DATA_FILES/merged_clean_CLUSTERS_GENES.csv") #this will create CSV file of clusters and genes                    
```

## Plot and View, name clusters 

```{r, warning=FALSE, message=FALSE}
DimPlot(object = merged_clean, reduction = "umap", label = TRUE, pt.size = 0.03)
FeaturePlot(merged_clean, features = c('EPCAM'), reduction = 'umap')

#TO NAME CLUSTERS:
Idents(object = merged_clean) <- 'UMAP_Clusters_new'
levels(merged_clean)
current.cluster.ids <- c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34)
new.cluster.ids <- c('Granulocyte','Granulocyte','Epi','Fibroblast','T-Cell','Exhausted CD8T','CD4T','Classical Macrophage','Epi','Endothelial','Endothelial','Endothelial','Cytotoxic CD8T','Epi','Alternatively Activated Macrophage','Fibroblast','Fibroblast','Epi','B-Cell','Fibroblast','Granulocyte','Pericyte','Erythrocytes','Smooth Muscle Cell','Pericyte','Fibroblast','Fibroblast','Epi','Nerves','Acinar','Alternatively Activated Macrophage','Mast ell','Nerves','Nerves','Epi')
names(x = new.cluster.ids) <- levels(x = merged_clean)
merged_clean <- RenameIdents(object = merged_clean, new.cluster.ids)

#Put these new cluster labels as metadata
merged_clean[["Level1Labels_Merged"]] <- Idents(object = merged_clean)
Idents(merged_clean) <- "Level1Labels_Merged"
levels(merged_clean)

#re-plot cluster with new cluster names so you can check that it worked:
merged_clean <- readRDS("../OBJECTS/Patient_1475/1475_merged_CORRECTED.rds")
DimPlot(object = merged_clean, reduction = "umap", label = TRUE, pt.size = 0.01)
DimPlot(object = merged_clean, split.by = 'time_point', reduction = "umap", label = TRUE, pt.size = 0.01)
#saveRDS(merged_clean, "../OBJECTS/Patient_1475/1475_merged_CORRECTED.rds")
``` 

# Now look at fibroblasts 

```{r}
#merged_clean <- readRDS("../OBJECTS/Patient_1475/1475_merged_CORRECTED.rds")
pt1475_fibros <- subset(merged_clean, idents = "Fibroblast")
pt1475_fibros <- NormalizeData(pt1475_fibros, normalization.method = "LogNormalize", scale.factor = 10000)
pt1475_fibros <- FindVariableFeatures(pt1475_fibros, selection.method = "vst", nfeatures = 2000, assay = "RNA")
pt1475_fibros <- ScaleData(pt1475_fibros, features = rownames(pt1475_fibros), verbose = FALSE)
pt1475_fibros <- RunPCA(pt1475_fibros, verbose = FALSE)
ElbowPlot(pt1475_fibros, 50)
PCNum = 15
pt1475_fibros <- FindNeighbors(pt1475_fibros, reduction = "pca", dims = 1:PCNum)
pt1475_fibros <- FindClusters(pt1475_fibros, resolution = 1.8)
pt1475_fibros <- RunUMAP(object = pt1475_fibros,dims = 1:PCNum)
DimPlot(object = pt1475_fibros, reduction = "umap", label = TRUE, pt.size = 0.03)
DimPlot(pt1475_fibros, reduction = "umap", label = TRUE, pt.size = 0.03, split.by = "time_point")

#save fibroblast object separate 
#saveRDS(pt1475_fibros, "../OBJECTS/Patient_1475/1475_T1_T2_T3_FIBROS.RDS")
#pt1475_fibros <- readRDS("../OBJECTS/Patient_1475/1475_T1_T2_T3_FIBROS.RDS")
DimPlot(pt1475_fibros, reduction = "umap", label = TRUE, pt.size = 0.03, split.by = "time_point")
```

## Plot Cell Frequencies 
```{r, warning=FALSE, message=FALSE}
#Let's make some histograms using the full object(not just fibroblasts)
#Now let's do a histogram by patient and time point
#splitting samples
pt1475 <- merged_clean
pt1475$Level1Labels_Merged
samples.list <- unique(pt1475$sample_id)
clusters <- lapply(samples.list, function(x){
  subset <- subset(pt1475, subset = sample_id == x)
  dist <- data.frame(table(subset$Level1Labels_Merged))
  
  return(dist)
})

names(clusters) <- samples.list

#calculate relative freq (fractions) of each cell type
clusters_percent <- lapply(clusters, FUN = function(x){
  summ <- sum(x$Freq)
  x$Freq <- (x$Freq/summ)
  return(x)
})

#making things ggplot-friendly!
clusters_dist <- reshape2::melt(clusters, id.var = "Var1")
colnames(clusters_dist) <- c("Level1Labels_Merged","variable","value","sample_id")   
clusters_percent_dist <- reshape2::melt(clusters_percent, id.var = "Var1")
colnames(clusters_percent_dist) <- c("Level1Labels_Merged","variable","value","sample_id")

ggplot(clusters_percent_dist, aes(fill=Level1Labels_Merged, y = value, x = sample_id)) + 
  scale_x_discrete(limits = c("1475T1", "1475T2", "1475 T3")) +
  geom_bar(position = "stack", stat = "identity") + 
  theme_bw() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, size = 8, face = "bold", hjust = 1)) +
  ylab("count") + xlab("patient_id") + ggtitle("Relative Cell Types Abundance") +
  scale_fill_manual(values = c('tan','wheat','darkseagreen1','darkslategray','darkorange','goldenrod3','darkorchid4','darkorange4','lightblue','gold','mediumorchid1','lavender','mediumpurple1','mediumvioletred','midnightblue','limegreen','violetred1'))
```





