---
title: "03_Merging_PT1475_Recurrence_Timepoint"
author: "Aylin Henstridge (Aylinhen@umich.edu)"
date: "2026-02-11"
output: 
  html_document:
    toc: true
    toc_depth: 2
    warnings: FALSE
    number_sections: true
    toc_float: true
    code_download: true
    fig_caption: true
    fig_width: 10
    fig_height: 6
    highlight: 'tango'
    theme: united
knit: (function(input, ...){rmarkdown::render(input, output_dir = "../DOCUMENTS")})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(toc = TRUE)
```

```{css, echo = FALSE}
pre {
  max-height: 500px;
  overflow-y: auto;
}

pre[class]{
  max-height: 500px;
}

.scroll-100{
  max-height: 500px;
  overflow-y: auto;
  background-color: inherit;
}
```

# MERGING 1475 T1, T2, T3
```{r, message=FALSE, warning=FALSE}
#################################################################
##                      loading libraries                      ##
############################################################################
library(Seurat)
library(SeuratObject)
library(parallel)
library(tidyverse)
library(ggplot2)
library(magrittr)
source("../UTILS/correct_reads_function.r")
library(dplyr)
library(data.table)
library(Matrix)
library(devtools)
library(RcppArmadillo)
library(Rcpp)
library(scales)
library(pheatmap)
library(gplots)
library(cowplot)
library(RColorBrewer)
library(viridis)
library(AUCell)
library(SummarizedExperiment)
`%nin%` <- Negate(`%in%`)
getPalette <- colorRampPalette(brewer.pal(9, 'Set1'))

#################################################################
##                        loading files                        ##
#################################################################

samples_info <- readxl::read_xlsx("../UTILS/sample_info_1475.xlsx")
```

We perform ambient RNA correction and merge like we did before but this time just 1475 and its 3 timepoints.
We use a different samples info sheet here. 

# Ambient RNA
```{r, message=FALSE, warning=FALSE, eval=FALSE}
#################################################################
##           Ambient RNA Correction (w +10% and Cap)           ##
#################################################################

## Ambient RNA correction was done using soupX were the contamination fraction
## is estimated from empty droplets and negative markers. An extra 10% was added
## as per the authors recommendation to remove 95-98% of the soup (ambient RNA).
## https://cran.r-project.org/web/packages/SoupX/vignettes/pbmcTutorial.html
## while capping the contamination fraction to 20%
##This is the same process we did when merging all T1 and T2 but now we do it with only 1475 T1,2, and 3.
samples_dir <- samples_info$cellranger7_output #list of my sample directories as seen in samples info
X = as.list(samples_dir)
samples_ID <- samples_info$sample_id 


# Set generous memory limit for future package (10GB per worker should be plenty)
options(future.globals.maxSize = 10000 * 1024^2)  # 10GB limit

corrected_objs <- mclapply(samples_ID, mc.cores = detectCores(), FUN = function(x){
  options(future.globals.maxSize = 10000 * 1024^2)
  message(paste0(x,"..."))
  sample <- samples_info %>% filter(sample_id == x)
  ctMtx <- sample$cellranger7_output
  seurat_obj <- correct_reads(data_dir = ctMtx, extra = 0.1, cap = TRUE)
  seurat_obj$sample_id <- sample$sample_id
  seurat_obj$timepoint <- sample$timepoint
  seurat_obj$patient_id <- sample$patient_id
  seurat_obj$tissue <- sample$tissue
  seurat_obj <- RenameCells(seurat_obj, new.names = paste0(colnames(seurat_obj),"_",unique(seurat_obj$sample_id)))
  
  seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-", assay = "RNA")
  seurat_obj[["percent.mt_CorrectedCounts"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-", assay = "CorrectedCounts")
  
  seurat_obj <- subset(seurat_obj, subset = nFeature_CorrectedCounts > 200 & percent.mt_CorrectedCounts < 15) #filter out cells and subset based on QC
  seurat_obj <- SCTransform(seurat_obj, vst.flavor = "v2", verbose = FALSE, conserve.memory = T, assay = 'CorrectedCounts') %>%
    RunPCA(verbose = FALSE) #Normalization 
  
  return(seurat_obj) 
  gc()
})

names(corrected_objs) <- samples_ID                      
saveRDS(corrected_objs, "../OBJECTS/1475_temp.rds")
```

# Merging the 3 timepoints now
You will use the 1475 temp file now like we did previously with the full object. 
```{r, message=FALSE, warning=FALSE, eval=FALSE}
###########################################################################
#              Generating Merged pt1475 Object                      #
############################################################################
tumor_obj <- readRDS("../OBJECTS/1475_temp.rds") #this is the list we generated above

#Samples integration
features <- SelectIntegrationFeatures(object.list = tumor_obj, nfeatures = 3000)
tumor_obj <- PrepSCTIntegration(object.list = tumor_obj, anchor.features = features)
anchors <- FindIntegrationAnchors(object.list = tumor_obj, normalization.method = "SCT",
                                  anchor.features = features)
tumor_obj <- IntegrateData(anchorset = anchors, normalization.method = "SCT") #this step takes quite a few hours, best to run over night 

saveRDS(tumor_obj, "../OBJECTS/1475_T123_merged.rds")
```

# Naming clusters

We will do this the same with feature plots, and can use the full Annotated object to help if needed. 

```{r, message=FALSE, warning=FALSE}
##########################################################################
#                                  Naming Clusters                      #
############################################################################

pt1475 <- readRDS("../OBJECTS/1475_T123_merged.rds")#read in merged 1475 object from above

DefaultAssay(pt1475) <- "integrated"
# dimension reduction and clustering of integrated object
set.seed(42)
pt1475 <- RunPCA(pt1475, verbose = FALSE)
pt1475 <- RunUMAP(pt1475, reduction = "pca", dims = 1:25, verbose = FALSE)
pt1475 <- FindNeighbors(pt1475, reduction = "pca", dims = 1:25)
pt1475 <- FindClusters(pt1475, resolution = 1.8)
pt1475[["seurat_clusters"]] <- Idents(object = pt1475)

#Look at what clusters look like at this step
DimPlot(object =pt1475, reduction = "umap", pt.size = 0.1, label = T, repel = T)

Idents(object = pt1475) <- 'seurat_clusters'
pt1475 <- PrepSCTFindMarkers(pt1475, assay = "SCT", verbose = TRUE)
markers <- FindAllMarkers(pt1475, assay = 'SCT', min.pct = 0.25, only.pos = T)
write.csv(markers, file = "../DATA/pt1475.markers.csv")
```

```{r, message=FALSE, warning=FALSE}
#Name the clusters, we have 47 clusters and will use common gene markers to identify them. 
DefaultAssay(pt1475) <- "SCT"
FeaturePlot(object = pt1475, features = c("EPCAM","KRT19", "MUC1", "KRT18"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order =T) #Epithelial
FeaturePlot(object = pt1475, features = c('PRSS1', 'AMY2A'), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order =T) #Acinar
FeaturePlot(object = pt1475, features = c("CDH5", "VWF"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01,  order =T) #Endothelial 
FeaturePlot(object = pt1475, features = c("PDGFRB", "PDGFRA", "PDPN", "ACTA2", "RGS5","PECAM1", "COL1A1", "IGFBP7"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #Fibroblasts and pericyte
FeaturePlot(object = pt1475, features = c("PTPRC", "CD14", "ITGAM", "MARCO", "APOE", "C1QA", "S100A8", "FCGR3A", "FCGR3B", "SPP1", "CD68", "HLA-DRA", 'G0S2'), cols = c("grey", "red"), reduction = "umap", pt.size = .05, order = T) #Myeloid
FeaturePlot(object = pt1475, features = c("CD3E", "CD4", "FOXP3", "CD8A", "GZMB", "PRF1", "NKG7", "NCAM1", "MKI67"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #T cells
FeaturePlot(object = pt1475, features = c("CD19", "CD79A", "IGJ", "IGLL5", "CD27", "CD138", 'MS4A1', 'JCHAIN'), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #B cells
FeaturePlot(object = pt1475, features = c("CLU", "TPSAB1", "CPA3"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #Mast cells
FeaturePlot(object = pt1475, features = c("MKI67"), cols = c("grey", "red"), reduction = "umap", pt.size = 0.01, order = T) #Cycling
FeaturePlot(object = pt1475, features = c('IRF7', 'GZMB', 'CLEC4C'), cols = c("grey", "red"), reduction = "umap", pt.size = .01, order = T) #pDCs
FeaturePlot(object = pt1475, features = c('PPBP', 'PF4', 'CLU'), cols = c("grey", "red"), reduction = "umap", pt.size = .01, order = T) #Platelet
FeaturePlot(object = pt1475, features = c('KRT20', 'CEACAM1', 'CA2', 'INS'), cols = c("grey", "red"), reduction = "umap", pt.size = .01, order = T) #Platelet

```


The featureplots above will tell us what each cluster is.
```{r, message=FALSE, warning=FALSE}
Idents(object = pt1475) <- 'seurat_clusters'
levels(pt1475)
current.cluster.ids <- c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14, 15, 16, 17, 18,
                         19, 20, 21, 22, 23, 24, 25,26, 27,28,29,30,31)
new.cluster.ids <- c('Epi', "Granulocyte", "Granulocyte", "Fibroblast", "CD8T", "Endothelial",
                     "CD8T","Fibroblast", "Granulocyte", "Epi", "CD8T",
                     "Macrophage", "Fibroblast", "CD4T", "Endothelial", "Acinar",
                     "Endothelial", "Granulocyte", "Pericyte", "B Cell", "Epi",
                     "Pericyte", "Epi", "Fibroblast", "Epi", "Macrophage",
                     "Fibroblast", "Nerve", "Epi", "Macrophage", "Mast", "Fibroblast")
names(x = new.cluster.ids) <- levels(x = pt1475)
pt1475<- RenameIdents(object = pt1475, new.cluster.ids)

DimPlot(object =pt1475, reduction = "umap", pt.size = 0.1, label = T, repel = T)
pt1475[["Level1Labels_Merged"]] <- Idents(object = pt1475)
Idents(object = pt1475) <- 'Level1Labels_Merged'
levels(pt1475)
```

```{r, eval=FALSE}
saveRDS(pt1475, "../OBJECTS/Annotated_pt1475_Object_AH.rds") 
```

# Cell Frequency Plot
```{r, message=FALSE, warning=FALSE}
#Let's make some histograms
Idents(pt1475) <- "sample_id"
levels(pt1475)
pt1475$sample_id[pt1475$sample_id == "6873-EC-1"] <- "1475_T2"
pt1475$sample_id[pt1475$sample_id == "4425-EC-1"] <- "1475_T1"
pt1475$sample_id[pt1475$sample_id == "9241-EC-1"] <- "1475_T3"

levels(pt1475)

Idents(pt1475) <- "Level1Labels_Merged"
pt1475$Level1Labels_Merged <- droplevels(pt1475@meta.data$Level1Labels_Merged)

#Now let's do a histogram by patient and time point
levels(pt1475)
Idents(pt1475) <- factor(Idents(pt1475), levels = rev(c('Epi', 'Acinar', 'Fibroblast', 'Pericyte', 'Endothelial', 'Macrophage', 'Granulocyte',  'Mast','CD8T',  'CD4T', 'Nerve', 
                                                                    'B Cell')))
#splitting samples
samples.list <- unique(pt1475$sample_id)
clusters <- lapply(samples.list, function(x){
  subset <- subset(pt1475, subset = sample_id == x)
  dist <- data.frame(table(subset$Level1Labels_Merged))
  
  return(dist)
})

names(clusters) <- samples.list

#calculate relative freq (fractions) of each cell type
clusters_percent <- lapply(clusters, FUN = function(x){
  summ <- sum(x$Freq)
  x$Freq <- (x$Freq/summ)
  return(x)
})
levels(pt1475)
#making things ggplot-friendly!
clusters_dist <- reshape2::melt(clusters, id.var = "Var1")
colnames(clusters_dist) <- c("Level1Labels_Merged","variable","value","sample_id")   
clusters_percent_dist <- reshape2::melt(clusters_percent, id.var = "Var1")
colnames(clusters_percent_dist) <- c("Level1Labels_Merged","variable","value","sample_id")

ggplot(clusters_percent_dist, aes(fill=Level1Labels_Merged, y = value, x = sample_id)) + 
  scale_x_discrete(limits = c('1475_T1', '1475_T2', '1475_T3')) +
  geom_bar(position = "stack", stat = "identity") + 
  theme_bw() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, size = 8, face = "bold", hjust = 1)) +
  ylab("count") + xlab("patient_id") + ggtitle("Relative Cell Types Abundance") +
  scale_fill_manual(values = c("wheat3", "peachpuff","aquamarine2", 'seagreen4','coral1', 'goldenrod', 'mediumorchid4', 'sienna3', 'lightskyblue1', 'yellow', 'magenta', 'lavender', 'thistle', 'violetred', 'dodgerblue3', 'chartreuse1'))

```

# Characterizing Fibroblasts Specifically 
```{r, message=FALSE, warning=FALSE}
#pick up here with fibroblasts
DefaultAssay(pt1475) <- "RNA"
pt1475 <- JoinLayers(pt1475)
DefaultAssay(pt1475) <- "CorrectedCounts"
pt1475_fibros <- subset(pt1475, idents = "Fibroblast")
pt1475 <- ScaleData(pt1475)
```

```{r, eval=FALSE}
#save fibroblast object separate 
saveRDS(pt1475_fibros, "../OBJECTS/1475_T1_T2_T3_FIBROS.RDS")
saveRDS(pt1475, "../OBJECTS/1475_merged_CORRECTED.rds")
```

```{r, message=FALSE, warning=FALSE}
##Fibroblast Heatmap Expression
fibro <- readRDS("../OBJECTS/1475_T1_T2_T3_FIBROS.RDS")
```


## Gene signatures
```{r, message=FALSE, warning=FALSE}
library(readxl)
mycaf_icaf <- read_excel("../GENE_SIGNATURES/MYCAF_ICAF.xlsx")
mycaf <- mycaf_icaf$mycaf
icaf <- mycaf_icaf$icaf
apcaf <- mycaf_icaf$apcaf

mycaf_icaf <- list(
  mycaf = mycaf,
  icaf = icaf,
  apcaf = apcaf
)
expr <- GetAssayData(fibro,
                     assay = "CorrectedCounts",
                     layer = "counts")

geneRanking <- AUCell_buildRankings(expr, nCores = 36, splitByBlocks = TRUE)

auc_score_2 <- AUCell_calcAUC(rankings = geneRanking, geneSets = mycaf_icaf, aucMaxRank = nrow(geneRanking)*0.1)

#make a data frame with ALL of the different types of auc scoring generated with different gene sets
auc_score_2 <- as.data.frame(t(assay(c(auc_score_2))))
auc_score_scaled_2 <- as.data.frame(scale(auc_score_2))
#scaled the data and add it as metadata
fibro <- AddMetaData(fibro, metadata = auc_score_scaled_2)

## Heatmap of AUCELL Scoring
tissue_cell_data <- FetchData(fibro, vars = c('mycaf','icaf','sample_id'))
n <- 1 # number of metadata columns
patient_avg <- data.frame()
for (id in levels(factor(tissue_cell_data$sample_id))) {
  data_subset <- tissue_cell_data %>% filter(sample_id == id)
  data_subset_avg <- apply(data_subset[,1:(ncol(data_subset)-n)], 2, mean)
  patient_avg <- rbind(patient_avg, data_subset_avg)
}
colnames(patient_avg) <- colnames(tissue_cell_data)[1:(ncol(tissue_cell_data)-n)]
rownames(patient_avg) <- levels(factor(tissue_cell_data$sample_id))
```

```{r, message=FALSE, warning=FALSE}
pheatmap(patient_avg, 
                    color = colorRampPalette(c("blue", "white", "red"))(256),  # Color palette
                    main = "Heatmap of Fibroblast Gene Sig.(scale by column)",   # Title of the heatmap
                    clustering_distance_rows = "euclidean",   # Distance metric for row clustering
                    clustering_distance_cols = "euclidean",   # Distance metric for column clustering
                    cellwidth = 10,
                    cellheight = 10,
                    fontsize = 8,           # Font size of row and column labels
                    legend = TRUE,           # Show legend
                    cluster_rows = FALSE,   # Do not cluster rows
                    cluster_cols = TRUE,    # Do not cluster columns
                    width = 100,   # Adjust the width of the plot
                    height = 100  # Adjust the height of the plot
)
```

# Adding the NR-CAF Signature 

```{r, message=FALSE, warning=FALSE}
#NR-CAF Signature
agf_signature <- c("NEGR1", "SLIT2", "ROBO1", "EPHA3","NRP1", "NFIA")
AGF <- list(
  
  agf_signature = agf_signature
)
expr <- GetAssayData(fibro,
                     assay = "CorrectedCounts",
                     layer = "counts")
geneRanking <- AUCell_buildRankings(expr, nCores = 36, splitByBlocks = TRUE)
auc_score <- AUCell_calcAUC(rankings = geneRanking, geneSets = AGF, aucMaxRank = nrow(geneRanking)*0.1)
#make a data frame with ALL of the different types of auc scoring generated with different gene sets
auc_score <- as.data.frame(t(assay(c(auc_score))))
auc_score_scaled <- as.data.frame(scale(auc_score))

#scaled the data and add it as metadata
fibro <- AddMetaData(fibro, metadata = auc_score_scaled)

## Define what "top expressors" means (e.g., top 10% of cells)
agf_threshold <- quantile(fibro$agf_signature, 0.75)  # top 25%
# 2. Create a new metadata column
fibro$agf <- ifelse(fibro$agf_signature >= agf_threshold, "agf", "caf")
# Check distribution
table(fibro$agf)
```

```{r, message=FALSE, warning=FALSE}
saveRDS(fibro, "../OBJECTS/agf_annotated_1475_CAFs.RDS")
```

# Final Heatmap
```{r, warning=FALSE, message=FALSE}
## Heatmap of AUCELL Scoring
tissue_cell_data <- FetchData(fibro, vars = c('mycaf','icaf','agf_signature', 'sample_id'))
n <- 1 # number of metadata columns
patient_avg <- data.frame()
for (id in levels(factor(tissue_cell_data$sample_id))) {
  data_subset <- tissue_cell_data %>% filter(sample_id == id)
  data_subset_avg <- apply(data_subset[,1:(ncol(data_subset)-n)], 2, mean)
  patient_avg <- rbind(patient_avg, data_subset_avg)
}
colnames(patient_avg) <- colnames(tissue_cell_data)[1:(ncol(tissue_cell_data)-n)]
rownames(patient_avg) <- levels(factor(tissue_cell_data$sample_id))
```

```{r, message=FALSE, warning=FALSE}
pheatmap(patient_avg, 
         color = colorRampPalette(c("blue", "white", "red"))(256),  # Color palette
         main = "Heatmap of Fibroblast Gene Sig.(scale by column)",   # Title of the heatmap
         clustering_distance_rows = "euclidean",   # Distance metric for row clustering
         clustering_distance_cols = "euclidean",   # Distance metric for column clustering
         cellwidth = 10,
         cellheight = 10,
         fontsize = 8,           # Font size of row and column labels
         legend = TRUE,           # Show legend
         cluster_rows = FALSE,   # Do not cluster rows
         cluster_cols = TRUE,    # Do not cluster columns
         width = 100,   # Adjust the width of the plot
         height = 100  # Adjust the height of the plot
)
```


