---
title: "05_T_Cell_Characterization"
author: "Aylin Henstridge (Aylinhen@umich.edu)"
date: "2025-12-09"
output: 
  html_document:
    toc: true
    toc_depth: 2
    warnings: FALSE
    number_sections: true
    toc_float: true
    code_download: true
    fig_caption: true
    fig_width: 10
    fig_height: 6
    highlight: 'tango'
    theme: united
knit: (function(input, ...){rmarkdown::render(input, output_dir = "../DOCUMENTS")})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(toc = TRUE)
```

```{css, echo = FALSE}
pre {
  max-height: 500px;
  overflow-y: auto;
}

pre[class]{
  max-height: 500px;
}

.scroll-100{
  max-height: 500px;
  overflow-y: auto;
  background-color: inherit;
}
```

# Introduction - Load Packages and Libraries and Objects (T-Cell)
```{r}
suppressPackageStartupMessages({
library(Seurat)
library(AUCell)
library(monocle3) #need this for the assay function in aucell script
library(GSVA)
library(pheatmap)
library(tidyverse)
library(RColorBrewer)
library(readxl)
library(DT)
library(fgsea)
library(msigdbr)
})
######################################################################################################################################
#read in subset object. this object contains the tcell cells for the dataset. 
tcell <- readRDS("../OBJECTS/tcell.RDS")
```


## plot full UMAP
```{r, warning=FALSE, message=FALSE}
Idents(tcell) <- "Level2Labels_T_Cell"
DimPlot(tcell, reduction = "umap", label = TRUE, pt.size = 0.03)
```

## Look at dot plot of t-cell gene expression
```{r, warning=FALSE, message=FALSE }
genes <- c("CD3E", "CD8A", "GZMB", "ZNF683", "GZMK", "EOMES", "CCR7", "SELL", "CD4", "FOXP3", "NKG7", "NCAM1")
DotPlot(object = tcell, features = genes, cols = c("white","magenta4")) + theme(legend.position= "right", axis.text.x = element_text(angle=45, size=10, hjust=1))
```


## Calculate relative freq (fractions) of each cell type
```{r, warning=FALSE, message=FALSE }
DefaultAssay(tcell) <- "CorrectedCounts"

tcell$Level2Labels_T_Cell
samples.list <- unique(tcell$sample_id)
clusters <- lapply(samples.list, function(x){
  subset <- subset(tcell, subset = sample_id == x)
  dist <- data.frame(table(subset$Level2Labels_T_Cell))
  
  return(dist)
})

names(clusters) <- samples.list

clusters_percent <- lapply(clusters, FUN = function(x){
  summ <- sum(x$Freq)
  x$Freq <- (x$Freq/summ)
  return(x)
})

#making things ggplot-friendly!
clusters_dist <- reshape2::melt(clusters, id.var = "Var1")
colnames(clusters_dist) <- c("Level2Labels_T_Cell","variable","value","sample_id")   
clusters_percent_dist <- reshape2::melt(clusters_percent, id.var = "Var1")
colnames(clusters_percent_dist) <- c("Level2Labels_T_Cell","variable","value","sample_id")

ggplot(clusters_percent_dist, aes(fill=Level2Labels_T_Cell, y = value, x = sample_id)) + 
  geom_bar(position = "stack", stat = "identity") + 
  theme_bw() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, size = 8, face = "bold", hjust = 1)) + scale_fill_manual(values = c('tan','wheat','darkseagreen1','darkslategray','darkorange','goldenrod3','darkorchid4','darkorange4','lightblue','gold','mediumorchid1','lavender','mediumpurple1','mediumvioletred','midnightblue','limegreen','violetred1'))
```

# Subset only T-Cells (no NK/NKT)

```{r, warning=FALSE, message=FALSE }
tcell <- subset(tcell, idents = c("Cytotoxic CD8T", "Exhausted CD8T", "CD4T", "TReg", "Naive CD8T"))

#make an interactive data table to search for genes present in clusters. 
tcell <- NormalizeData(tcell, normalization.method = "LogNormalize", scale.factor = 10000)
tcell <- FindVariableFeatures(tcell, selection.method = "vst", nfeatures = 2000, assay = "CorrectedCounts")
tcell <- ScaleData(tcell)
markers <- FindAllMarkers(tcell, assay = "CorrectedCounts", min.pct = 0.25, only.pos = T) 
datatable((markers), filter = 'top', caption = "Seurat Cluster Markers", options = list(scrollX = T))
```

# Differentially Expressed Genes 
```{r, warning=FALSE, message=FALSE }
DefaultAssay(object = tcell) <- "CorrectedCounts" #make sure you always use corrected counts when finding markers
Idents(object = tcell) <- 'time_point'
tcell <- ScaleData(tcell, features = rownames(tcell))
tcell.markers <- FindMarkers(tcell, ident.1 = "1", ident.2 = "2", assay = "CorrectedCounts", test.use = "MAST", verbose = FALSE, recorrect_umi = FALSE, logfc.threshold = 0)
#the verbose, prepsctfindmarkers, SCT assay, and recorrect umi are all only included if SCT assay is used. you can leave off for corrected counts. 
#write.csv(tcell.markers, file = "../DATA_FILES/tcell_scaled_DE_2025.csv")
datatable((tcell.markers), filter = 'top', caption = "tcell Scaled DE Markers", options = list(scrollX = T))

#saveRDS(tcell, "../OBJECTS/tcell_only.RDS")
```

```{r, warning=FALSE, message=FALSE }
Idents(tcell) <- "Level2Labels_T_Cell"
DimPlot(tcell, split.by = "time_point")
```

# Subset CD4 v. CD8 v. TREG
```{r, warning=FALSE, message=FALSE }
#######tcell.markers######################################################################################################################################

#Subset CD4 T Cells
Idents(tcell) <- "Level2Labels_T_Cell"
levels(tcell)

CD4 <- subset(tcell, idents = c("CD4T")) 
CD8 <- subset(tcell, idents = c("Cytotoxic CD8T", "Exhausted CD8T", "Naive CD8T"))
Treg <- subset(tcell, idents = c("TReg"))
```

# SLOPE PLOTS
```{r, warning=FALSE, message=FALSE }
#TIGIT , CD44, CXCR4, CXCL12 (in CD4 & CD8 TCELLS SEPERATELY)

genes <- c('TIGIT', 'CD44', 'CXCR4', 'CXCL12')
expr_data_CD4 <- FetchData(CD4, vars = genes)
expr_data_CD4$patient_id <- CD4@meta.data$patient_id
expr_data_CD4$time_point <- CD4@meta.data$time_point
avg_expr <- expr_data_CD4 %>%
  group_by(patient_id, time_point) %>%
  summarise(across(all_of(genes), mean, na.rm = TRUE), .groups = "drop")

avg_expr_long <- avg_expr %>%
  pivot_longer(cols = all_of(genes), names_to = "gene", values_to = "avg_expr")

ggplot(avg_expr_long, aes(x = time_point, y = avg_expr, group = patient_id, color = patient_id)) +
  geom_line() +
  geom_point() +
  facet_wrap(~gene, scales = "free_y") +
  theme_minimal() +
  labs(title = "Average Gene Expression Before and After Treatment",
       y = "Average Expression", x = "Time_Point")

expr_data_CD8 <- FetchData(CD8, vars = genes)
expr_data_CD8$patient_id <- CD8@meta.data$patient_id
expr_data_CD8$time_point <- CD8@meta.data$time_point
avg_expr <- expr_data_CD8 %>%
  group_by(patient_id, time_point) %>%
  summarise(across(all_of(genes), mean, na.rm = TRUE), .groups = "drop")

avg_expr_long <- avg_expr %>%
  pivot_longer(cols = all_of(genes), names_to = "gene", values_to = "avg_expr")

ggplot(avg_expr_long, aes(x = time_point, y = avg_expr, group = patient_id, color = patient_id)) +
  geom_line() +
  geom_point() +
  facet_wrap(~gene, scales = "free_y") +
  theme_minimal() +
  labs(title = "Average Gene Expression Before and After Treatment",
       y = "Average Expression", x = "Time_Point")

#######################################################################################################################
```

## CXCR4 is increased post-treatment in all t-cells, further analysis of these cells. 
```{r, warning=FALSE, message=FALSE }
FeaturePlot(tcell, features = c("CXCR4"), split.by = "time_point", pt.size = 2,) & scale_color_gradientn(colours =rev(brewer.pal(n=11, name = "RdYlBu"))) & theme(legend.position = "right")  # Ensures the color bar is visible
#Isolating CXCR4 High v. Low population 
tcellprintdata <- FetchData(object = tcell, vars = c("CXCR4"))
# find the lowest and highest neuro quartile

quantile(tcellprintdata$CXCR4)
#0%      25%      50%      75%     100% 
#0.000000 1.153924 2.004530 2.688180 4.649832 

#Then you are going to which cells your metadata with a threshold based on 50% quartile
CXCR4_low <- WhichCells(object = tcell, expression = CXCR4 < 2.688180)
CXCR4_hi <- WhichCells(object = tcell, expression = CXCR4 > 2.688180)

Idents(tcell, cells = CXCR4_hi) <- 'CXCR4_hi'
Idents(tcell, cells = CXCR4_low) <- 'CXCR4_low'
tcell$CXCR4_Stat <- Idents(tcell)
levels(tcell)
```

## Plotting CXCR4 Expression
```{r, warning=FALSE, message=FALSE }
Idents(tcell) <- "CXCR4_Stat"
DimPlot(object = tcell, reduction = "umap", pt.size = 1, label = F)
DimPlot(object = tcell, split.by = "time_point",reduction = "umap", pt.size = 0.01, label = F, cols = c("gray", "red"))
```

## CXCR4 differentially expressed genes
```{r, warning=FALSE, message=FALSE }
Idents(tcell) <- "CXCR4_Stat"
CXCR4.markers <- FindMarkers(tcell, ident.1 = "CXCR4_hi", ident.2 = "CXCR4_low", assay = "CorrectedCounts", test.use = "MAST", verbose = FALSE, recorrect_umi = FALSE, logfc.threshold = 0)

```

# Markers of exhasuation and activation 
```{r}
###############################################################################################################################
Inhibitory  <- c("BTLA", 'CTLA4', 'HAVCR2', 'PDCD1', 'TIGIT', 'TNFRSF18')
Stimulatory <- c('CD27', 'CD28', 'CD40LG', 'ICOS', 'TNFRSF4', 'TNFRSF9')

expr_data_CD4 <- FetchData(CD4, vars = Inhibitory)
expr_data_CD4$patient_id <- CD4@meta.data$patient_id
expr_data_CD4$time_point <- CD4@meta.data$time_point
avg_expr <- expr_data_CD4 %>%
  group_by(patient_id, time_point) %>%
  summarise(across(all_of(Inhibitory), mean, na.rm = TRUE), .groups = "drop")

avg_expr_long <- avg_expr %>%
  pivot_longer(cols = all_of(Inhibitory), names_to = "gene", values_to = "avg_expr")

ggplot(avg_expr_long, aes(x = time_point, y = avg_expr, group = patient_id, color = patient_id)) +
  geom_line() +
  geom_point() +
  facet_wrap(~gene, scales = "free_y") +
  theme_minimal() +
  labs(title = "Average Inhibitory Gene Expression Before and After Treatment",
       y = "Average Expression", x = "Time_Point")
#################################################################################################################
```

```{r, warning=FALSE, message=FALSE }
expr_data_CD4 <- FetchData(CD4, vars = Stimulatory)
expr_data_CD4$patient_id <- CD4@meta.data$patient_id
expr_data_CD4$time_point <- CD4@meta.data$time_point
avg_expr <- expr_data_CD4 %>%
  group_by(patient_id, time_point) %>%
  summarise(across(all_of(Stimulatory), mean, na.rm = TRUE), .groups = "drop")

avg_expr_long <- avg_expr %>%
  pivot_longer(cols = all_of(Stimulatory), names_to = "gene", values_to = "avg_expr")

ggplot(avg_expr_long, aes(x = time_point, y = avg_expr, group = patient_id, color = patient_id)) +
  geom_line() +
  geom_point() +
  facet_wrap(~gene, scales = "free_y") +
  theme_minimal() +
  labs(title = "Average Stimulatory Gene Expression Before and After Treatment",
       y = "Average Expression", x = "Time_Point")
```

```{r, warning=FALSE, message=FALSE }
#####################################################################################################
expr_data_CD8 <- FetchData(CD8, vars = Inhibitory)
expr_data_CD8$patient_id <- CD8@meta.data$patient_id
expr_data_CD8$time_point <- CD8@meta.data$time_point
avg_expr <- expr_data_CD8 %>%
  group_by(patient_id, time_point) %>%
  summarise(across(all_of(Inhibitory), mean, na.rm = TRUE), .groups = "drop")

avg_expr_long <- avg_expr %>%
  pivot_longer(cols = all_of(Inhibitory), names_to = "gene", values_to = "avg_expr")

ggplot(avg_expr_long, aes(x = time_point, y = avg_expr, group = patient_id, color = patient_id)) +
  geom_line() +
  geom_point() +
  facet_wrap(~gene, scales = "free_y") +
  theme_minimal() +
  labs(title = "Average Gene Expression Before and After Treatment",
       y = "Average Expression", x = "Time_Point")
```

```{r, warning=FALSE, message=FALSE }
######################################################################################################################################
expr_data_CD8 <- FetchData(CD8, vars = Stimulatory)
expr_data_CD8$patient_id <- CD8@meta.data$patient_id
expr_data_CD8$time_point <- CD8@meta.data$time_point
avg_expr <- expr_data_CD8 %>%
  group_by(patient_id, time_point) %>%
  summarise(across(all_of(Stimulatory), mean, na.rm = TRUE), .groups = "drop")

avg_expr_long <- avg_expr %>%
  pivot_longer(cols = all_of(Stimulatory), names_to = "gene", values_to = "avg_expr")

ggplot(avg_expr_long, aes(x = time_point, y = avg_expr, group = patient_id, color = patient_id)) +
  geom_line() +
  geom_point() +
  facet_wrap(~gene, scales = "free_y") +
  theme_minimal() +
  labs(title = "Average Gene Expression Before and After Treatment",
       y = "Average Expression", x = "Time_Point")



######################################################################################################################################
```

# PSEUDOBULK CD4, CD8
```{r, warning=FALSE, message=FALSE }
source("../UTILS/pseudobulk_functions.R")
##################################################################
##                Pseudobulk for each population                ##
##################################################################
CD4_DGE <- pseudobulk_dge(seurat_object = CD4, assay = "CorrectedCounts",
                          aggregate_by = "sample_id", comparison = "time_point")
CD8_DGE <- pseudobulk_dge(seurat_object = CD8, assay = "CorrectedCounts",
                          aggregate_by = "sample_id", comparison = "time_point")
```

## PCA Plots
```{r, warning=FALSE, message=FALSE }
##############################################################
##                          PCA Plots                          ##
#################################################################

CD4_Average_expression <- AverageExpression(CD4, group.by = c("patient_id"))

CD4_PLOT <- pca_plot(CD4_DGE, groups = "time_point") + ggtitle("CD4 T-Cell Population")
CD8_PLOT <- pca_plot(CD8_DGE, groups = "time_point") + ggtitle("CD8 T-Cell Population")

CD4_PLOT
CD8_PLOT


################################################################################################
```

# Violin Plots of Differentially Expressed Genes & CAF Interactions

```{r, warning=FALSE, message=FALSE }
library(ggpubr)
Idents(CD4) <- "time_point"
Idents(CD8) <- "time_point"
VlnPlot(CD4, features = c("CXCR4"), pt.size = 0, cols = c("pink", "magenta")) + stat_compare_means(method = "wilcox.test")
VlnPlot(CD8, features = c("CXCR4"), pt.size = 0, cols = c("pink", "magenta"))+ stat_compare_means(method = "wilcox.test")
```

```{r, warning=FALSE, message=FALSE }
cafs <- readRDS("../OBJECTS/final_fibroblasts_for_part2.RDS")  
Idents(cafs) <- "time_point"
VlnPlot(cafs, features = c('CXCL12'), pt.size = 0, cols = c("pink", "magenta"))   
VlnPlot(cafs, features = c('CXCR4'), pt.size = 0, cols = c("pink", "magenta"))
```

