---
title: "07_Fibroblast_Cell_Characterization"
author: "Aylin Henstridge (Aylinhen@umich.edu)"
date: "2026-02-11"
output: 
  html_document:
    toc: true
    toc_depth: 2
    warnings: FALSE
    number_sections: true
    toc_float: true
    code_download: true
    fig_caption: true
    fig_width: 10
    fig_height: 6
    highlight: 'tango'
    theme: united
knit: (function(input, ...){rmarkdown::render(input, output_dir = "../DOCUMENTS")})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(toc = TRUE)
```

```{css, echo = FALSE}
pre {
  max-height: 500px;
  overflow-y: auto;
}

pre[class]{
  max-height: 500px;
}

.scroll-100{
  max-height: 500px;
  overflow-y: auto;
  background-color: inherit;
}
```

# Load in libraries and data
```{r, message=FALSE, warning=FALSE}
##fibroblast signatures

##Aylin henstridge 3/13/2024
library(Seurat)
library(SeuratObject)
library(SeuratWrappers)
library(AUCell)
library(monocle3)
library(GSVA)
library(pheatmap)
library(tidyverse)
library(RColorBrewer)
library(DT)
library(readxl)
library(EnhancedVolcano)
library(enrichplot)
library(cowplot)
library(DESeq2)
library(ggalt)
library(devtools)
library(ashr)
library(tidytree)
library(dplyr)
library(fgsea)
library(msigdbr)
library(ggpubr)
`%nin%` <- Negate(`%in%`)
getPalette <- colorRampPalette(brewer.pal(9, 'Set1'))
source("../UTILS/pseudobulk_functions.R")
source("../UTILS/patient_split.R")
source("../UTILS/GSEA.R")
#read in subsested object. this object contains the fibroblasts/fibro only epithelial cells for the dataset. 

#FibroEndo subset
FibroEndo<- readRDS("../OBJECTS/Fibro_subset.rds")
DimPlot(object =FibroEndo, reduction = "umap", pt.size = 0.1, label = F, repel = T, cols = getPalette(length(levels(FibroEndo))), split.by= "timepoint")
```

# Clean up object
```{r, message=FALSE, warning=FALSE}
DefaultAssay(FibroEndo) <- "integrated"
# dimension reduction and clustering of integrated object
FibroEndo <- RunPCA(FibroEndo, verbose = FALSE)
ElbowPlot(FibroEndo, 50)
FibroEndo <- RunUMAP(FibroEndo, reduction = "pca", dims = 1:30, verbose = FALSE)
FibroEndo <- FindNeighbors(FibroEndo, reduction = "pca", dims = 1:30)
FibroEndo <- FindClusters(FibroEndo, resolution = 1.0)
FibroEndo[["FibroEndo_seurat_clusters"]] <- Idents(object = FibroEndo)
```

```{r, message=FALSE, warning=FALSE}
DimPlot(object =FibroEndo, reduction = "umap", pt.size = 0.1, label = T, repel = T, cols = getPalette(length(levels(FibroEndo))))
```

```{r, message=FALSE, warning=FALSE}
DefaultAssay(FibroEndo) <- "SCT"

FibroEndo <- PrepSCTFindMarkers(FibroEndo, assay = "SCT", verbose = TRUE)
markers <- FindAllMarkers(FibroEndo, assay = 'SCT', min.pct = 0.25, only.pos = T)

#Name the clusters level 1
Idents(object = FibroEndo) <- 'FibroEndo_seurat_clusters'
levels(FibroEndo)
current.cluster.ids <- c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14, 15)
DimPlot(FibroEndo, label = T)
new.cluster.ids <- c('Fibro2', 'Endothelial', 'Fibro2', 'Fibro2', 'Fibro1', 'Fibro2', 
                     'Fibro3', 'Fibro1', 'Fibro3', 'Fibro1', 'Endothelial',
                     'Fibro3', 'Fibro2', 'Endothelial', 'Junk', 'Fibro1')
names(x = new.cluster.ids) <- levels(x = FibroEndo)
FibroEndo<- RenameIdents(object = FibroEndo, new.cluster.ids)

FibroEndo[["Level1Labels_FibroEndo"]] <- Idents(object = FibroEndo)
FibroEndo_fil <- subset(FibroEndo, idents = c('Junk'), invert = TRUE)
Idents(object=FibroEndo_fil) <- 'Level1Labels_FibroEndo'
Idents(FibroEndo_fil) <- factor(Idents(FibroEndo_fil), levels = rev(c('Fibro1', 'Fibro2', 'Fibro3', 'Endothelial')))
levels(FibroEndo_fil)
DefaultAssay(FibroEndo_fil) <- "SCT"
```

# Label top expressed genes in DotPlot

```{r, message=FALSE, warning=FALSE}
DotPlot(object = FibroEndo_fil, features = c("SLIT2", 'NEGR1', 'RORA', 'EGFR', 'COL1A1', 'COL1A2', 'SPARC', "DPT", 
                                             'LGALS2', 'ACTA2', 'NOTCH3', 'TAGLN', 'RGS5', 'PLVAP', 'VWF', 
                                             'PECAM1'), cols = c("white", "magenta4")) +theme(legend.position = "right", 
                                                                                              axis.text.x = element_text(angle = 45, size = 10,  hjust = 1))
```

```{r, message=FALSE, warning=FALSE}
#Have contaminating pericytes present in Fibro3 - these need to be removed
DimPlot(object =FibroEndo_fil, reduction = "umap", pt.size = 0.1, label = F, repel = T, split.by= "timepoint")
```
```{r, eval=FALSE}
saveRDS(FibroEndo_fil, "Annotated_FibroEndoPeri_AH.rds")
```

```{r, message=FALSE, warning=FALSE}
#######################Now filter out endothelial cells and characterize fibroblasts specifically#####################################

Idents(object = FibroEndo_fil) <- "Level1Labels_FibroEndo"

#make an interactive data table to search for genes present in clusters. 
markers <- FindAllMarkers(FibroEndo_fil, assay = "CorrectedCounts", min.pct = 0.25, only.pos = T) 
datatable((markers), filter = 'top', caption = "Seurat Cluster Markers", options = list(scrollX = T))
```

```{r, message=FALSE, warning=FALSE}
#change identity to what you want cluster sorted by in UMAP (i.e cluster names here)
DimPlot(FibroEndo_fil, reduction = "umap", label = TRUE, pt.size = 0.03)
```

```{r, message=FALSE, warning=FALSE}
Idents(FibroEndo_fil) <- "Level1Labels_FibroEndo"
FeaturePlot(FibroEndo_fil, features = c("RGS5"))

fibro <- subset(FibroEndo_fil, idents = c("Fibro1", "Fibro2")) #Fibro3 is actually pericytes, so remove them
```

# Re-Cluster Clean Object

```{r, message=FALSE, warning=FALSE}
######################################Now recluster just fibroblasts###################################################
DefaultAssay(fibro) <- "integrated"
fibro <- FindVariableFeatures(fibro, selection.method = "vst", nfeatures = 2000, assay = "CorrectedCounts")
fibro <- ScaleData(fibro, features = rownames(fibro),verbose = FALSE)
fibro <- RunPCA(fibro, verbose = FALSE)
ElbowPlot(fibro, 50)
PCNUM = 15
fibro <- RunUMAP(fibro, reduction = "pca", dims = 1:PCNUM)
fibro <- FindNeighbors(fibro, reduction = "pca", dims = 1:PCNUM)
fibro <- FindClusters(fibro, resolution = 0.17)
levels(fibro)
current.cluster.ids <- c('0', '1', '2')
new.cluster.ids <- c('Fibro1', 'Fibro2', 'Fibro3')
names(x = new.cluster.ids) <- levels(x = fibro)
fibro<- RenameIdents(object = fibro, new.cluster.ids)
fibro[["Level1Labels_Fibro_only"]] <- Idents(object = fibro)
```

```{r, message=FALSE, warning=FALSE}
DimPlot(object = fibro, reduction = "umap", label = FALSE
        , pt.size = 0.5, cols = c("salmon", "chartreuse4", "cornflowerblue"))
```

# Cell_frequency plot

```{r, message=FALSE, warning=FALSE}
#check gene expression in fibro 1,2,3
DefaultAssay(fibro) <- "CorrectedCounts"
fibro <- NormalizeData(fibro)
fibro <- ScaleData(fibro, features = rownames(fibro),verbose = FALSE)
VlnPlot(fibro,features = c("FAP", "PDGFRA", "PDGFRB"), pt.size = 0)
VlnPlot(fibro, features = c("ACTA2", "TAGLN", "POSTN"), pt.size = 0)
VlnPlot(fibro, features = c("PDPN", "DPT", "MFAP4"), pt.size = 0)
VlnPlot(fibro, features = c("CD34", "RSPO3", "CDH11"), pt.size = 0)
```

```{r, message=FALSE, warning=FALSE}
fibro$Level1Labels_Fibro_only
samples.list <- unique(fibro$sample_id)
clusters <- lapply(samples.list, function(x){
  subset <- subset(fibro, subset = sample_id == x)
  dist <- data.frame(table(subset$Level1Labels_Fibro_only))
  
  return(dist)
})

names(clusters) <- samples.list

#calculate relative freq (fractions) of each cell type
clusters_percent <- lapply(clusters, FUN = function(x){
  summ <- sum(x$Freq)
  x$Freq <- (x$Freq/summ)
  return(x)
})

#making things ggplot-friendly!
clusters_dist <- reshape2::melt(clusters, id.var = "Var1")
colnames(clusters_dist) <- c("Level1Labels_Fibro_only","variable","value","sample_id")   
clusters_percent_dist <- reshape2::melt(clusters_percent, id.var = "Var1")
colnames(clusters_percent_dist) <- c("Level1Labels_Fibro_only","variable","value","sample_id")

ggplot(clusters_percent_dist, aes(fill=Level1Labels_Fibro_only, y = value, x = sample_id)) + 
  geom_bar(position = "stack", stat = "identity") + 
  theme_bw() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, size = 8, face = "bold", hjust = 1)) +
  ylab("count") + xlab("patient_id") + ggtitle("Relative Cell Types Abundance") +
  scale_fill_manual(values = c('salmon', 'chartreuse4', 'cornflowerblue'))
```

```{r, message=FALSE, warning=FALSE}
genes <- c("PDGFRA", "NEGR1", "SLIT2", "ROBO1", 
           "SEMA3A", "NRP1", "NFIA",
           "EPHA3", "FAP",
           "ACTA2", "TAGLN", "THY1",
           "COL1A1", "COL1A2",
           "VIM", "PDPN",
           "POSTN", "CXCL12", "DPT")
```

```{r, message=FALSE, warning=FALSE}
DotPlot(object = fibro, features = genes, cols = c("white","magenta4"), scale = FALSE) + theme(legend.position= "right", axis.text.x = element_text(angle=45, size=10, hjust=1))
DotPlot(
  object = fibro,
  features = genes,
  scale = TRUE
) +
  scale_color_gradientn(
    colours = rev(brewer.pal(11, "RdYlBu"))
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, size = 10, hjust = 1)
  )
```

# Differentially Expressed Genes
```{r, message=FALSE, warning=FALSE}
DefaultAssay(fibro) <- "CorrectedCounts"
fibro.markers <- FindAllMarkers(fibro, only.pos = TRUE, min.pct = 0.25,logfc.threshold = 0.25) #always use 0.25 for percent and fold change threshold
fibro.markers %>% group_by(cluster) %>% top_n(n=2, wt=avg_log2FC) #ranks biggest to smallest
```                 

# AUCELL ANALYSIS ON FIBROBLASTS USING KNOWN GENE SIGNATURES
```{r, message=FALSE, warning=FALSE}
library(readxl)
mycaf_icaf <- read_excel("../GENE_SIGNATURES/MYCAF_ICAF.xlsx")
mycaf <- mycaf_icaf$mycaf
icaf <- mycaf_icaf$icaf
apcaf <- mycaf_icaf$apcaf


mycaf_icaf <- list(
  mycaf = mycaf,
  icaf = icaf,
  apcaf = apcaf
)
expr <- GetAssayData(fibro,
                     assay = "CorrectedCounts",
                     layer = "counts")
geneRanking <- AUCell_buildRankings(expr, nCores = 36, splitByBlocks = TRUE)

auc_score_2 <- AUCell_calcAUC(rankings = geneRanking, geneSets = mycaf_icaf, aucMaxRank = nrow(geneRanking)*0.1)

#make a data frame with ALL of the different types of auc scoring generated with different gene sets
auc_score_2 <- as.data.frame(t(assay(c(auc_score_2))))
auc_score_scaled_2 <- as.data.frame(scale(auc_score_2))
#scaled the data and add it as metadata
fibro <- AddMetaData(fibro, metadata = auc_score_scaled_2)
```

## Create the plot object and display it in the console
```{r, message=FALSE, warning=FALSE}
FeaturePlot(fibro, features = c("mycaf"), label = FALSE, repel = TRUE, order = TRUE) & 
  scale_colour_gradientn(scale = TRUE, colours = rev(brewer.pal(n = 11, name = "RdYlBu")))
FeaturePlot(fibro, features = c("icaf"), label = FALSE, repel = TRUE, order = TRUE) & 
  scale_colour_gradientn(scale = TRUE, colours = rev(brewer.pal(n = 11, name = "RdYlBu")))
FeaturePlot(fibro, features = c("apcaf"), label = FALSE, repel = TRUE, order = TRUE) & 
  scale_colour_gradientn(scale = TRUE, colours = rev(brewer.pal(n = 11, name = "RdYlBu")))
```

## Feature plot of top expressed genes
```{r, message=FALSE, warning=FALSE}

###___________________________________________________
FeaturePlot(fibro, features = c("NEGR1"), pt.size = 1)& scale_color_gradientn(colours =rev(brewer.pal(n=11, name = "RdYlBu"))) & theme(legend.position = "right")  # Ensures the color bar is visible
FeaturePlot(fibro, features = c("NFIA"), pt.size = 1) & scale_color_gradientn(colours =rev(brewer.pal(n=11, name = "RdYlBu"))) & theme(legend.position = "right")  # Ensures the color bar is visible
FeaturePlot(fibro, features = c("ROBO1"), pt.size = 1) & scale_color_gradientn(colours =rev(brewer.pal(n=11, name = "RdYlBu"))) & theme(legend.position = "right")  # Ensures the color bar is visible
FeaturePlot(fibro, features = c("NRP1"), pt.size = 1)& scale_color_gradientn(colours =rev(brewer.pal(n=11, name = "RdYlBu"))) & theme(legend.position = "right")  # Ensures the color bar is visible
FeaturePlot(fibro, features = c("SLIT2"), pt.size = 1) & scale_color_gradientn(colours =rev(brewer.pal(n=11, name = "RdYlBu"))) & theme(legend.position = "right")  # Ensures the color bar is visible
FeaturePlot(fibro, features = c("ACTA2"), pt.size = 1) & scale_color_gradientn(colours =rev(brewer.pal(n=11, name = "RdYlBu"))) & theme(legend.position = "right")  # Ensures the color bar is visible
FeaturePlot(fibro, features = c("CXCL12"), pt.size = 1)& scale_color_gradientn(colours =rev(brewer.pal(n=11, name = "RdYlBu"))) & theme(legend.position = "right")  # Ensures the color bar is visible
FeaturePlot(fibro, features = c("EPHA3"), pt.size = 1) & scale_color_gradientn(colours =rev(brewer.pal(n=11, name = "RdYlBu"))) & theme(legend.position = "right")  # Ensures the color bar is visible
```


## Make cleaner timepoint metadata
```{r, message=FALSE, warning=FALSE}
fibro$time_point <- NA
Idents(fibro) <- "time_point"
levels(fibro)
fibro$time_point[fibro$sample_id == "1402_T1"] <- 1
fibro$time_point[fibro$sample_id == "1402_T2"] <- 2
fibro$time_point[fibro$sample_id == "1475_T1"] <- 1
fibro$time_point[fibro$sample_id == "1475_T2"] <- 2
fibro$time_point[fibro$sample_id == "1529_T1"] <- 1
fibro$time_point[fibro$sample_id == "1529_T2"] <- 2
fibro$time_point[fibro$sample_id == "1550_T1"] <- 1
fibro$time_point[fibro$sample_id == "1550_T2"] <- 2
fibro$time_point[fibro$sample_id == "1471_T1"] <- 1
fibro$time_point[fibro$sample_id == "1471_T2"] <- 2

Idents(fibro) <- "time_point"
fibro$time_point <- factor(
  fibro$time_point,
  levels = c(1, 2)
)

Idents(fibro) <- "time_point"
levels(fibro)
```

# Look at differential expressed genes
```{r, message=FALSE, warning=FALSE}
Idents(fibro) <- "patient_id"
levels(fibro)
Idents(fibro) <- factor(Idents(fibro), levels = rev(c('1402', '1471', '1475', '1529', '1550')))
VlnPlot(fibro, features = c("NEGR1"), split.by = "time_point", split.plot = TRUE, pt.size = 0)
VlnPlot(fibro, features = c("NRP1"), split.by = "time_point", split.plot = TRUE, pt.size = 0)
VlnPlot(fibro, features = c("NFIA"), split.by = "time_point", split.plot = TRUE, pt.size = 0)
VlnPlot(fibro, features = c("EPHA3"), split.by = "time_point", split.plot = TRUE, pt.size = 0)
VlnPlot(fibro, features = c("SLIT2"), split.by = "time_point", split.plot = TRUE, pt.size = 0)
VlnPlot(fibro, features = c("ACTA2"), split.by = "time_point", split.plot = TRUE, pt.size = 0)
VlnPlot(fibro, features = c("DPT"), split.by = "time_point", split.plot = TRUE, pt.size = 0)
VlnPlot(fibro, features = c("LRRC15"), split.by = "time_point", split.plot = TRUE, pt.size = 0)
VlnPlot(fibro, features = c("FAP"), split.by = "time_point", split.plot = TRUE, pt.size = 0)
VlnPlot(fibro, features = c("CXCL12"), split.by = "time_point", split.plot = TRUE, pt.size = 0)
```

# Defining Neural CAF signature
```{r, message=FALSE, warning=FALSE}
Idents(fibro) <- 'Level1Labels_Fibro_only'
DimPlot(fibro)
```

## DE code:(using corrected counts assay) 
```{r, message=FALSE, warning=FALSE}
DefaultAssay(object = fibro) <- "CorrectedCounts" #make sure you always use corrected counts when finding markers
Fibro_scaled <- ScaleData(fibro, features = rownames(fibro))
Idents(object = Fibro_scaled) <- 'time_point'
levels(Fibro_scaled)
```

## Neural Gene Signature
```{r, message=FALSE, warning=FALSE}
agf_signature <- c("NEGR1", "SLIT2", "ROBO1", "EPHA3","NRP1", "NFIA")
AGF <- list(
  
  agf_signature = agf_signature
)
expr <- GetAssayData(fibro,
                     assay = "CorrectedCounts",
                     layer = "counts")

geneRanking <- AUCell_buildRankings(expr, nCores = 36, splitByBlocks = TRUE)
auc_score <- AUCell_calcAUC(rankings = geneRanking, geneSets = AGF, aucMaxRank = nrow(geneRanking)*0.1)
#make a data frame with ALL of the different types of auc scoring generated with different gene sets
auc_score <- as.data.frame(t(assay(c(auc_score))))
auc_score_scaled <- as.data.frame(scale(auc_score))

#scaled the data and add it as metadata
fibro <- AddMetaData(fibro, metadata = auc_score_scaled)
```

```{r, message=FALSE, warning=FALSE}
FeaturePlot(fibro, features = c("agf_signature"), pt.size =1, label = FALSE, repel = TRUE) & 
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlBu")))
```


```{r, message=FALSE, warning=FALSE}
# 1. Define what "top expressors" means (e.g., top 10% of cells)
agf_threshold <- quantile(fibro$agf_signature, 0.75)  # top 25%
# 2. Create a new metadata column
fibro$agf <- ifelse(fibro$agf_signature >= agf_threshold, "agf", "caf")
# Check distribution
table(fibro$agf)
```

# Differential Expression and Pathway Analysis
```{r, message=FALSE, warning=FALSE}
Idents(fibro) <- "agf"
levels(fibro)

agf.markers <- FindMarkers(fibro, ident.1 = "agf", ident.2 = "caf", assay = "CorrectedCounts", test.use = "MAST", verbose = FALSE, recorrect_umi = FALSE, logfc.threshold = 0)
```

## Published Gene Sig. and Neural Sig.

```{r, message=FALSE, warning=FALSE}
tissue_cell_data <- FetchData(fibro, vars = c('agf_signature','icaf','mycaf', 'sample_id'))
n <- 1 # number of metadata columns
patient_avg <- data.frame()
for (id in levels(factor(tissue_cell_data$sample_id))) {
  data_subset <- tissue_cell_data %>% filter(sample_id == id)
  data_subset_avg <- apply(data_subset[,1:(ncol(data_subset)-n)], 2, mean)
  patient_avg <- rbind(patient_avg, data_subset_avg)
}
colnames(patient_avg) <- colnames(tissue_cell_data)[1:(ncol(tissue_cell_data)-n)]
rownames(patient_avg) <- levels(factor(tissue_cell_data$sample_id))
cycling <- pheatmap(patient_avg, 
                    color = colorRampPalette(c("blue", "white", "red"))(256),
                    scale = "column",# Color palette,
                    main = "Heatmap of Fibroblast Gene Sig.(scale by column)",   # Title of the heatmap
                    clustering_distance_rows = "euclidean",   # Distance metric for row clustering
                    clustering_distance_cols = "euclidean",   # Distance metric for column clustering
                    cellwidth = 10,
                    cellheight = 10,
                    fontsize = 8,           # Font size of row and column labels
                    legend = TRUE,           # Show legend
                    cluster_rows = FALSE,   # Do not cluster rows
                    cluster_cols = FALSE,    # Do not cluster columns
                    width = 100,   # Adjust the width of the plot
                    height = 100,   # Adjust the height of the plot
                    xlab = "signatures",  # X-axis label
                    ylab = "Patient_Timepoint"   # Y-axis labelhttp://127.0.0.1:22419/graphics/9f4d525b-e106-4a69-8ae6-695b611e88a4.png
)
```

## Published CAF Genes
```{r, message=FALSE, warning=FALSE}
DefaultAssay(fibro) <- "CorrectedCounts"
Idents(fibro) <- "patient_id"
VlnPlot(fibro, features = c("LRRC15", "DPT", "PDGFRB", "PDGFRA",
                            "TAGLN", "FAP", "PDPN"), pt.size = 0, split.by = "time_point", split.plot = TRUE)
VlnPlot(fibro, features = c("LEF1", "FABP4"), pt.size = 0, split.by = "time_point", split.plot = TRUE)
```

