---
title: "11. Merging External Datasets and Integrating Pt. 1"
author: "Aylin Henstridge (Aylinhen@umich.edu)"
date: "2025-12-09"
output: 
  html_document:
    toc: true
    toc_depth: 2
    warnings: FALSE
    number_sections: true
    toc_float: true
    code_download: true
    fig_caption: true
    fig_width: 10
    fig_height: 6
    highlight: 'tango'
    theme: united
knit: (function(input, ...){rmarkdown::render(input, output_dir = "../DOCUMENTS")})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(toc = TRUE)
```

```{css, echo = FALSE}
pre {
  max-height: 500px;
  overflow-y: auto;
}

pre[class]{
  max-height: 500px;
}

.scroll-100{
  max-height: 500px;
  overflow-y: auto;
  background-color: inherit;
}
```

# Load object and libraries

```{r, warning=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(DT)
  library(cowplot)
  library(patchwork)
  library(AUCell)
})

longitduinal <- readRDS("../OBJECTS/eileen_annotated_longitudinal.rds")
gift_of_life_full<- readRDS("../OBJECTS/integrated_data_realigned_hg38.rds")
```

# Prepare Data and Set Correct Metadata
```{r, eval = FALSE}
DefaultAssay(longitduinal) <- "CorrectedCounts"
DefaultAssay(gift_of_life_full) <- "CorrectedCounts"

gift_of_life_full$sample_id <- gift_of_life_full$sample_name

gift_of_life_full$Level1Labels <- gift_of_life_full$cell_types
Idents(gift_of_life_full) <- "Level1Labels"
```

# Merge Objects
```{r, eval =FALSE, warning=FALSE}
merged_object <- merge(longitduinal, gift_of_life_full)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#saveRDS(merged_object, "../OBJECTS/merged_longitduinal_GoL.rds")

merged_object <- readRDS("../OBJECTS/merged_longitduinal_GoL.rds")
```

# Inetgrate Objects and Pass Labels
```{r, warning=FALSE, eval=FALSE}
DefaultAssay(merged_object) <- 'CorrectedCounts'

samples.list <- SplitObject(merged_object, split.by = "sample_id")
samples.list <- lapply(X = samples.list, FUN = function(x) {
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

rm(merged_object)
gc()
```
## remove samples with less than 100 cells
```{r, eval = FALSE}
samples.list <- samples.list[-3]

features <- SelectIntegrationFeatures(object.list = samples.list)
samples.list <- lapply(X = samples.list, FUN = function(x) {
  x <- ScaleData(x, features = features, verbose = FALSE)
  x <- RunPCA(x, features = features, verbose = FALSE)
})

anchors <- FindIntegrationAnchors(object.list = samples.list, anchor.features = features, reduction = "rpca",assay = rep("CorrectedCounts", length(samples.list)))

Integrated_Dataset <- IntegrateData(anchorset = anchors)
```

## Normalize and Cluster
```{r, eval=FALSE}
DefaultAssay(Integrated_Dataset) <- "CorrectedCounts"
Integrated_Dataset <- FindVariableFeatures(Integrated_Dataset, selection.method = "vst", nfeatures = 2000)
Integrated_Dataset <- ScaleData(Integrated_Dataset, verbose = FALSE)
Integrated_Dataset <- RunPCA(Integrated_Dataset, verbose = FALSE)
Integrated_Dataset <- RunUMAP(Integrated_Dataset, reduction = "pca", dims = 1:20, reduction.name = 'unintegrated_umap')

DefaultAssay(Integrated_Dataset) <- "integrated"
Integrated_Dataset <- ScaleData(Integrated_Dataset, verbose = FALSE)
Integrated_Dataset <- RunPCA(Integrated_Dataset, verbose = FALSE)
Integrated_Dataset <- RunUMAP(Integrated_Dataset, reduction = "pca", dims = 1:25, reduction.name = 'integrated_umap')
Integrated_Dataset <- FindNeighbors(Integrated_Dataset, reduction = "pca", dims = 1:25)
Integrated_Dataset <- FindClusters(Integrated_Dataset, resolution = 1.8)

Integrated_Dataset[["UMAP_Clusters"]] <- Idents(object = Integrated_Dataset)  
#saveRDS(Integrated_Dataset, "../OBJECTS/integrated_merged_datasets_longitduinal_hg38_GOL.rds")
```

```{r, echo = FALSE, warning=FALSE, message=FALSE}
Integrated_Dataset <- readRDS("../OBJECTS/integrated_merged_datasets_longitduinal_hg38_GOL.rds")
```

```{r, warning=FALSE, message=FALSE}
Idents(Integrated_Dataset) <- 'sample_id'
levels(Integrated_Dataset)
p1 <- DimPlot(Integrated_Dataset,  reduction = "integrated_umap",  group.by = c("sample_id"),) + NoLegend()
p2 <- DimPlot(Integrated_Dataset,  reduction = "unintegrated_umap",  group.by = c("sample_id"))
p1+p2

Idents(Integrated_Dataset) <- "Level1Labels"
levels(Integrated_Dataset)
p1 <- DimPlot(Integrated_Dataset,  reduction = "integrated_umap",  group.by = c("Level1Labels")) + NoLegend()  
p2 <- DimPlot(Integrated_Dataset,  reduction = "unintegrated_umap",  group.by = c("Level1Labels"))
p1+p2

Idents(Integrated_Dataset) <- "UMAP_Clusters"
DimPlot(Integrated_Dataset, reduction = "integrated_umap", label = TRUE)
```

```{r, eval = FALSE}
markers <- FindAllMarkers(Integrated_Dataset, assay = "CorrectedCounts", min.pct = 0.25, only.pos = T) 

#write.csv(markers, "../DATA_FILES/MERGED_GOL_FULL_CLUSTER_MARKER_LIST.csv")
```

```{r, warning=FALSE, message=FALSE}
# TO NAME CLUSTERS:
Idents(object = Integrated_Dataset) <- 'UMAP_Clusters'
levels(Integrated_Dataset)
current.cluster.ids <- c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,43,44,45,46,47,48,49,50,51,52,53,54,55,56,57,58,59,60,61,62)
new.cluster.ids <- c('CD4T', 'CD8T', 'Acinar', 'Granulocyte', 'Epi', 'Granulocyte', 'Epi', 'Acinar', 'B-Cell', 'Epi', 'Macrophage', 'Epi', 'Macrophage', 'Fibroblast', 'CD4T', 'CD8T', 'Pericyte', 'Granulocyte', 'CD8T', 'Granulocyte', 'NK', 'Macrophage', 'Fibroblast', 'Acinar', 'Macrophage', 'Macrophage', 'Acinar', 'Mast', 'Cycling', 'Endothelial', 'Fibroblast', 'Endothelial', 'Epi', 'Acinar', 'Plasma','Granulocyte', 'Macrophage', 'Macrophage', 'Acinar', 'Acinar','Fibroblast','Macrophage', 'Acinar', "Epi", 'Macrophage', 'Acinar', 'TReg', 'Endothelial', 'Epi', 'Macrophage', 'Epi', 'Epi', 'Epi', 'Pericyte', 'Endothelial', 'Plasma', 'NK', 'Epi', 'Epi', 'Plasma', 'Platelet', 'Granulocyte', 'CD8T')
names(x = new.cluster.ids) <- levels(x = Integrated_Dataset)
Integrated_Dataset <- RenameIdents(object = Integrated_Dataset, new.cluster.ids)
Integrated_Dataset[["Merged_Level1Labels"]] <- Idents(object = Integrated_Dataset)
```

## Create new metadata now
```{r, warning=FALSE, message=FALSE}
Idents(Integrated_Dataset) <- 'sample_id'
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '3079-EC-2'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '6894-EC-1'] <- "Treated Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '4358-EC-1'] <- "Treated Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '5392-EC-1'] <- "Treated Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '4425-EC-1'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '6873-EC-1'] <- "Treated Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '5161-EC-1'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '6216-EC-1'] <- "Treated Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '5570-EC-1'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '6846-EC-1'] <- "Treated Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '3829-EC'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '3861-EC'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '3862-EC'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '4160-EC'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '4161-EC'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '4346-EC'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '4347-EC'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '4637-EC-1'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '4637-EC-2'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '4741-EC-1'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '4741-EC-2'] <- "Healthy"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '119833'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '125796'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '135041'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '125421'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '126197'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '123798'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '127260'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '130289'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '130450'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '130840'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '130742'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '135040'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '135156'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '137333'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '135334'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '135714'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '136365'] <- "Treatment Naive Tumor"
Integrated_Dataset$DiseaseState[Integrated_Dataset$sample_id == '136687'] <- "Treatment Naive Tumor"
Idents(Integrated_Dataset) <- "DiseaseState"
#saveRDS(Integrated_Dataset, "Integrated_Final_GOL_Longitudinal_Merge.rds")
```

# Subset Fibroblasts
```{r, warning=FALSE, message=FALSE}
Idents(Integrated_Dataset) <- "Merged_Level1Labels"
FIBROS <- subset(Integrated_Dataset, idents = c('Fibroblast'))

#Integrated Data set is everything
#tumor_only is tumor only fibroblasts
#FIBROS is all fibroblasts including healthy
```

```{r, warning=FALSE, message=FALSE}
DefaultAssay(FIBROS) <- 'CorrectedCounts'
DimPlot(FIBROS, reduction = "integrated_umap")

Idents(FIBROS) <- "DiseaseState"
levels(FIBROS)
tumor_only <- subset(FIBROS, idents = c("Treatment Naive Tumor", "Treated Tumor"))
Idents(tumor_only) <- "DiseaseState"
levels(tumor_only)
```

## Compare overall treatment naive tumor and treated tumor. 
```{r, warning=FALSE, message=FALSE}
library(ggpubr)
VlnPlot(tumor_only, feature = c("ACTA2"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(tumor_only, feature = c("NEGR1"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(tumor_only, feature = c("NFIA"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(tumor_only, feature = c("NRP1"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(tumor_only, feature = c("SLIT2"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(tumor_only, feature = c("EPHA3"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
```

## Create new metadata to break it down to just tumor and just healthy regardless of treatment status
```{r, warning=FALSE, message=FALSE}
FIBROS$tumor_type <- ifelse(
  FIBROS$DiseaseState == "Healthy", 
  "Healthy", 
  ifelse(FIBROS$DiseaseState %in% c("Treatment Naive Tumor", "Treated Tumor"), "PDAC", NA)
)

Idents(FIBROS) <- "tumor_type"
levels(FIBROS)
FIBROS$tumor_type <- factor(FIBROS$tumor_type, levels = c("Healthy", "PDAC"))
FIBROS <- subset(FIBROS, idents = c("PDAC", "Healthy"))
#now compare healthy & tumor in full object
library(ggpubr)
VlnPlot(FIBROS, feature = c("ACTA2"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("NEGR1"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("NFIA"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("NRP1"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("SLIT2"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("EPHA3"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("TAGLN"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("DPT"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("LRRC15"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("FAP"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("CXCL12"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("PDGFRA"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("PDGFRB"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("PDPN"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, feature = c("LEF1"), pt.size = 0) + stat_compare_means(method = "wilcox.test")
```

## look at NR-CAF signature in full object
```{r, warning=FALSE, message=FALSE}
FeaturePlot(FIBROS, reduction = "integrated_umap", features = c("NEGR1"))
FeaturePlot(FIBROS, reduction = "integrated_umap", features = c("NRP1"))
FeaturePlot(FIBROS, reduction = "integrated_umap", features = c("ACTA2"))
FeaturePlot(FIBROS, reduction = "integrated_umap", features = c("EPHA3"))
FeaturePlot(FIBROS, reduction = "integrated_umap", features = c("SEMA3A"))
FeaturePlot(FIBROS, reduction = "integrated_umap", features = c("ROBO1"))
FeaturePlot(FIBROS, reduction = "integrated_umap", features = c("NFIA"))

#########################################################################################
```

## Apply NR-CAF signature to CAFs only *tumor_only*, apply them to FIBROS too
```{r, warning=FALSE, message=FALSE}
AGF <- c("NEGR1", "NRP1", "NFIA", "EPHA3", "SLIT2", "SEMA3A", "ROBO1")

library(readxl)
library(AUCell)

mycaf_icaf <- read_excel("../GENE_SIGNATURES//MYCAF_ICAF.xlsx")
mycaf <- mycaf_icaf$mycaf
icaf <- mycaf_icaf$icaf
apcaf <- mycaf_icaf$apcaf

AGF <- c("NEGR1", "SLIT2", "ROBO1", "NFIA", "NRP1", "SEMA3A", "EPHA3")
AGF_metadata <- list(
  AGF = AGF
)

mycaf_icaf <- list(
  
  mycaf = mycaf,
  apcaf = apcaf,
  icaf = icaf
)

expr <- GetAssayData(tumor_only,
                     assay = "CorrectedCounts",
                     layer = "counts")


expr2 <- GetAssayData(FIBROS,
                     assay = "CorrectedCounts",
                     layer = "counts")
agf <- list(
  
  agf = AGF
)
library(SummarizedExperiment)
library(RColorBrewer)
geneRanking <- AUCell_buildRankings(expr, nCores = 36, splitByBlocks = TRUE)
geneRanking2 <- AUCell_buildRankings(expr2, nCores = 36, splitByBlocks = TRUE)

auc_score <- AUCell_calcAUC(rankings = geneRanking, geneSets = agf, aucMaxRank = nrow(geneRanking)*0.1)
auc_score_2 <- AUCell_calcAUC(rankings = geneRanking, geneSets = mycaf_icaf, aucMaxRank = nrow(geneRanking)*0.1)
auc_score <- as.data.frame(t(assay(c(auc_score))))
auc_score_scaled <- as.data.frame(scale(auc_score))
auc_score_2 <- as.data.frame(t(assay(c(auc_score_2))))
auc_score_scaled_2 <- as.data.frame(scale(auc_score_2))

auc_score_3 <- AUCell_calcAUC(rankings = geneRanking2, geneSets = agf, aucMaxRank = nrow(geneRanking2)*0.1)
auc_score_4 <- AUCell_calcAUC(rankings = geneRanking2, geneSets = mycaf_icaf, aucMaxRank = nrow(geneRanking2)*0.1)
auc_score_3 <- as.data.frame(t(assay(c(auc_score_3))))
auc_score_scaled_3 <- as.data.frame(scale(auc_score_3))
auc_score_4 <- as.data.frame(t(assay(c(auc_score_4))))
auc_score_scaled_4 <- as.data.frame(scale(auc_score_4))

tumor_only <- AddMetaData(tumor_only, metadata = auc_score_scaled)
tumor_only <- AddMetaData(tumor_only, metadata = auc_score_scaled_2)

FIBROS <- AddMetaData(FIBROS, metadata = auc_score_scaled_3)
FIBROS <- AddMetaData(FIBROS, metadata = auc_score_scaled_4)
```

## Create the plot object and display it in the console
```{r, warning=FALSE, message=FALSE}
#check number of cells
table(tumor_only@meta.data$orig.ident)
table(FIBROS@meta.data$orig.ident)
plot <- FeaturePlot(tumor_only, features = c("mycaf", "icaf", "apcaf", "agf"), reduction = "integrated_umap", label = FALSE, repel = TRUE) & 
scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlBu")))
print(plot)

DimPlot(FIBROS, reduction = "integrated_umap", split.by = 'DiseaseState')
FeaturePlot(FIBROS, features = c('NEGR1'), split.by = "DiseaseState")        
FeaturePlot(FIBROS, features = c('ACTA2'), split.by = "DiseaseState")   
```

# Remove healthy patients
```{r, warning=FALSE, message=FALSE}
agf_threshold <- quantile(FIBROS$agf, 0.75)  # top 25%
# 2. Create a new metadata column
FIBROS$agf <- ifelse(FIBROS$agf >= agf_threshold, "agCAF", "CAF")
# Check distribution
ag_caf_cafs <- table(FIBROS$agf, FIBROS$patient_id)
print(ag_caf_cafs)
```

# Pass NR-CAF labels to object
```{r, warning=FALSE, message=FALSE}
FIBROS$passed_level1labels <- FIBROS$agf
Idents(FIBROS) <- "passed_level1labels"
levels(FIBROS)
#now we created metadata with all of the cell types 
Idents(Integrated_Dataset) <- "Merged_Level1Labels" 
levels(Integrated_Dataset)
Integrated_Dataset_nofibro <- subset(Integrated_Dataset, idents = c("Fibroblast"), invert = TRUE)
Idents(Integrated_Dataset_nofibro) <- "Merged_Level1Labels"
Integrated_Dataset_nofibro$passed_level1labels <- Integrated_Dataset_nofibro$Merged_Level1Labels
#passing all labels to main integrated data set 
passed_level1labels <- rbind(Integrated_Dataset_nofibro@meta.data %>% select(passed_level1labels), FIBROS@meta.data %>% select(passed_level1labels))
aligned_labels <- passed_level1labels[rownames(Integrated_Dataset@meta.data), ]
Integrated_Dataset@meta.data$passed_level1labels <- aligned_labels
Idents(Integrated_Dataset) <- "passed_level1labels"

#changing colors of the umap
cols <- c(
  "CD8T"        = "aquamarine2", 
  "CD4T"        = "coral1",
  "TReg"        = "sienna3",
  "Cycling"     = "violetred",
  "Pericyte"    = "lavender",
  "Platelet"    = "brown",
  "Macrophage"  = "peachpuff", 
  "Mast"        = "chartreuse1", 
  "NK"          = "seagreen4", 
  "B-Cell"      = "lightskyblue1",   # fixed here
  "Plasma"      = "dodgerblue3", 
  "CAF"         = "goldenrod", 
  "Endothelial" = "pink", 
  "Epi"         = "mediumorchid4",
  "Granulocyte" = "wheat3",
  "Acinar"      = "thistle",
  "agCAF"       = "magenta")
```

```{r, warning=FALSE, message=FALSE}
DimPlot(Integrated_Dataset, reduction = "integrated_umap", label = F, cols = cols)
```

# Removing healthy tissue again
```{r, warning=FALSE, message=FALSE}
##############Removing Healthy Tissue from Merged Database - Tumor Only#############################
Idents(Integrated_Dataset) <- "DiseaseState"
cancer_patients <- subset(Integrated_Dataset, idents = c("Treatment Naive Tumor", "Treated Tumor"))
Idents(cancer_patients) <- "passed_level1labels"
levels(cancer_patients)
DimPlot(cancer_patients, reduction = "integrated_umap", label = F, cols = cols)
table(cancer_patients@meta.data$orig.ident)
```

# Looking at expression of Ligand-Receptor Pairs
```{r, warning=FALSE, message=FALSE}
##############Identifying common LR Pairs#############################
Idents(cancer_patients) <- "passed_level1labels"

CELLS_INTEREST <- subset(cancer_patients, idents = c("Acinar", "Platelet", "Cycling", "Mast"), invert=TRUE)

DefaultAssay(CELLS_INTEREST) <- "CorrectedCounts"
NRP1 <- c("NRP1", "VEGFA",
           "SEMA3B", "SEMA3A", "SEMA3C", "PLXND1")

nc_WNT <- c("WNT5A", "WNT11",
         "FZD2", "FZD5", "FZD7",
         "ROR1", "ROR2", "RYK")

c_WNT <- c("WNT2", "WNT2B","WNT7A", "WNT3A", 
           "WNT7B", "WNT9A", "WNT10A",
           "FZD2", "FZD4", "FZD5", "FZD7", "FZD1", "FZD8", "LRP5", "LRP6")

hedgehog <- c("SHH", "DHH", "IHH", "PTCH1", "PTCH2")

NEGR1 <- c("NEGR1", "ADAM10", "FGFR2"
           )
library(RColorBrewer)
DotPlot(object = CELLS_INTEREST, features = NEGR1, scale = TRUE) + theme(
    legend.position = "right",
    axis.text.x = element_text(angle=45, size=10, hjust=1)
  ) + scale_color_gradientn(
    colours = rev(brewer.pal(n = 11, name = "RdYlBu"))
  )

```

# Differential Expression and Pathway Analysis
```{r, warning=FALSE, message=FALSE}
Idents(cancer_patients) <- "passed_level1labels"
FIBROS <- subset(cancer_patients, ideant = c("CAF", "agCAF"))
Idents(FIBROS) <- "passed_level1labels"
levels(FIBROS)

agf.markers <- FindMarkers(FIBROS, ident.1 = "agCAF", ident.2 = "CAF", assay = "RNA", test.use = "MAST", verbose = FALSE, recorrect_umi = FALSE, logfc.threshold = 0)
```

## KEGG Reactome
```{r, eval = FALSE, warning=FALSE, message=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)

up_in_t1 <- agf.markers %>%
  filter(avg_log2FC > 0 )

down_in_t1 <- agf.markers %>%
  filter(avg_log2FC < 0)

up_ids <- bitr(up_in_t1$gene, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)
down_ids <- bitr(down_in_t1$gene, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)

up_kegg <- enrichKEGG(gene=up_ids$ENTREZID, organism="hsa", pvalueCutoff=0.05)
down_kegg <- enrichKEGG(gene=down_ids$ENTREZID, organism="hsa", pvalueCutoff=0.05)

up_df <- as.data.frame(up_kegg)
down_df <- as.data.frame(down_kegg)

up_df$direction <- "Down in T1"
down_df$direction <- "Up in T1"

up_df$enrichment_score <- -log10(up_df$p.adjust)
down_df$enrichment_score <- log10(down_df$p.adjust)

combined_df <- bind_rows(up_df, down_df) %>%
  filter(!is.infinite(enrichment_score)) %>%
  arrange(desc(abs(enrichment_score))) %>%
  slice(1:50)  # adjust number of categories shown

# Plot
ggplot(combined_df, aes(x = reorder(Description, enrichment_score), y = enrichment_score, fill = direction)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("Up in T1" = "blue", "Down in T1" = "red")) +
  labs(title = "KEGG Pathways Enriched in T3",
       x = "Pathway", y = "Signed -log10(adjusted p-value)") +
  theme_minimal(base_size = 14)
```

# Look at gene expression of signaling pathways 
## Wnt
```{r, warning=FALSE, message=FALSE, eval=FALSE}

Idents(FIBROS) <- "agf"
VlnPlot(FIBROS, features = c("WNT5A", "SFRP1", "MYC", "ROR2", "PRICKLE1", 
                             "FZD7", "PLCB4", "LRP6"), pt.size = 0)

VlnPlot(FIBROS, features = c("WNT5A"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("SFRP1"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("MYC"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("ROR2"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("PRICKLE1"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("FZD7"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("PLCB4"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("LRP6"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")

Idents(FIBROS) <- "DiseaseState"
FIBROS <- subset(FIBROS, idents = c("Treatment Naive Tumor", "Treated Tumor"))
VlnPlot(FIBROS, features = c("WNT5A", "SFRP1", "MYC",
                             "PRICKLE1", 
                             "FZD7", "PLCB4", "LRP6"), pt.size = 0)
VlnPlot(FIBROS, features = c("WNT5A"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("SFRP1"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("MYC"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("PRICKLE1"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("FZD7"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("PLCB4"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")
VlnPlot(FIBROS, features = c("LRP6"), pt.size = 0)+ stat_compare_means(method = "wilcox.test")

```













