---
title: "13_Spatial_Transcriptomics_Part_2"
author: "Aylin Henstridge (Aylinhen@umich.edu)"
date: "2026-06-24"
output: 
  html_document:
    toc: true
    toc_depth: 2
    warnings: FALSE
    number_sections: true
    toc_float: true
    code_download: true
    fig_caption: true
    fig_width: 10
    fig_height: 6
    highlight: 'tango'
    theme: united
knit: (function(input, ...){rmarkdown::render(input, output_dir = "../DOCUMENTS")})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(toc = TRUE)
```

```{css, echo = FALSE}
pre {
  max-height: 500px;
  overflow-y: auto;
}

pre[class]{
  max-height: 500px;
}

.scroll-100{
  max-height: 500px;
  overflow-y: auto;
  background-color: inherit;
}
```

# Load in libraries and data
```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
library(Seurat)
library(SeuratObject)
library(SeuratWrappers)
library(AUCell)
library(monocle3)
library(GSVA)
library(pheatmap)
library(tidyverse)
library(RColorBrewer)
library(DT)
library(dplyr)
`%nin%` <- Negate(`%in%`)
getPalette <- colorRampPalette(brewer.pal(9, 'Set1'))
source("../UTILS/pseudobulk_functions.R")
})
```

# Open Spatial Object (Pei et al, 2025, Nature)
```{r}
PDAC_combined <- readRDS("../OBJECTS/PDAC_combined_cytospace_deconvolution.rds") #read in deconvoluted object
#object was shared with us from published data  from DOI: 10.1038/s41586-025-08927-x
```

## Subset cell types of interest from the deconvoluted object
We want to subset out CAFs and epithelial cells from the object.

```{r, results='hide', warning=FALSE}
PDAC_combined$celltype_cytospace <- ifelse(PDAC_combined$CAFS >= 0.5, "CAF",
                                 ifelse(PDAC_combined$Epicells >= 0.5, "Epi", "Other"))

Idents(PDAC_combined) <- "celltype_cytospace"
levels(PDAC_combined) #We now have new metadata that says we have CAFs and EPi cells
table(PDAC_combined$celltype_cytospace) #and it looks like we have 13376 CAFs and 3196 epithelial cells 

caf_epi_cytospace <- subset(PDAC_combined, idents = c("CAF", "Epi"))
```

```{r, echo=FALSE}
DimPlot(object =PDAC_combined, reduction = "umap", pt.size = 0.05, raster = F, label = T, repel = T, cols = getPalette(length(levels(PDAC_combined)))) 
DimPlot(object =caf_epi_cytospace, reduction = "umap", pt.size = 0.05, raster = F, label = T, repel = T, cols = getPalette(length(levels(PDAC_combined)))) 
```

### SUBSET CAFS

We will update CAF metadata and then map agCAF gene expression in the different sites. 

```{r, results='hide', warning=FALSE}
CAFs <- subset(caf_epi_cytospace, idents = c("CAF"))
DimPlot(object =CAFs, reduction = "umap", pt.size = 0.05, raster = F, label = T, repel = T, cols = getPalette(length(levels(PDAC_combined)))) 

DefaultAssay(CAFs) <- "Spatial" 
CAFs$site[CAFs$site == "PC"] <- "Peritoneal Metastases"
CAFs$site[CAFs$site == "Lung"] <- "Lung Metastases"
CAFs$site[CAFs$site == "Liver"] <- "Liver Metastases"
CAFs$site[CAFs$site == "Panc"] <- "Primary Tumor"


genes <- c("PDGFRA", "NEGR1", "SLIT2", "ROBO1", 
           "SEMA3A", "NRP1", "NFIA",
           "EPHA3", "FAP",
           "ACTA2", "CXCL12")
Idents(CAFs) <- "site"
```

```{r, echo=FALSE}
DotPlot(
  object = CAFs,
  features = genes,
  scale = TRUE
) +
  scale_color_gradientn(
    colours = rev(brewer.pal(11, "RdYlBu"))
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, size = 10, hjust = 1)
  )
```

#### AUCELL ANALYSIS ON CAFs USING KNOWN GENE SIGNATURES
```{r, results='hide', warning=FALSE}
##AUCELL ANALYSIS ON CAFsBLASTS USING KNOWN GENE SIGNATURES
library(readxl)
mycaf_icaf <- read_excel("../GENE_SIGNATURES/MYCAF_ICAF.xlsx")
mycaf <- mycaf_icaf$mycaf
icaf <- mycaf_icaf$icaf
apcaf <- mycaf_icaf$apcaf

###___________________________________________________
agf_signature <- c("NEGR1", "SLIT2", "ROBO1", "EPHA3","NRP1", "NFIA", 'SEMA3A')

sigantures <- list(
  mycaf = mycaf,
  icaf = icaf,
  apcaf = apcaf,
  agf_signature = agf_signature
)
expr <- GetAssayData(CAFs,
                     assay = "Spatial",
                     layer = "counts")
geneRanking <- AUCell_buildRankings(expr, nCores = 36, splitByBlocks = TRUE)
auc_score_2 <- AUCell_calcAUC(rankings = geneRanking, geneSets = sigantures, aucMaxRank = nrow(geneRanking)*0.1)

#do not scale aucel score for this so it stays comparable across different groups 
#make a data frame with ALL of the different types of auc scoring generated with different gene sets
auc_score_2 <- as.data.frame(t(assay(c(auc_score_2))))
auc_score_scaled_2 <- as.data.frame(auc_score_2)
#add it as metadata
CAFs <- AddMetaData(CAFs, metadata = auc_score_scaled_2)
```

```{r}
FeaturePlot(CAFs, features = c("mycaf", "icaf", "agf_signature"), pt.size =1, label = FALSE, repel = TRUE, order = TRUE) & 
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlBu")))
```

```{r, warning=FALSE}
# 1. Define what threshold for  "top expressors" of each score
agf_threshold <- quantile(CAFs$agf_signature)  # top 25% (neural caf signature)
agf_threshold

mycaf_thresh <- quantile(CAFs$mycaf)  # top 25%
mycaf_thresh

icaf_thresh <- quantile(CAFs$icaf)  # top 25%
icaf_thresh

library(dplyr)
#Now we will define neural CAFs,myCAfs, iCAFs, and indeterminate or hybrid CAF types  
CAFs$double_positives <- case_when(
  CAFs$myCAF <  0.53 & CAFs$agf_signature >= 0.059 & CAFs$icaf >= 0.13  ~ "agf_icaf",
  CAFs$myCAF >= 0.53 & CAFs$agf_signature <  0.059 & CAFs$icaf <  0.13  ~ "mycaf",
  CAFs$myCAF <  0.53 & CAFs$agf_signature >= 0.059 & CAFs$icaf <  0.13  ~ "agf",
  CAFs$myCAF <  0.53 & CAFs$agf_signature <  0.059 & CAFs$icaf >= 0.13  ~ "icaf",
  TRUE ~ "undefined_caf"
)

table(CAFs$double_positives, useNA = "ifany")

PDAC_combined$double_positives <- "Other Cells"
PDAC_combined$double_positives <- CAFs$double_positives

PDAC_combined$double_positives <- ifelse( PDAC_combined$celltype_cytospace == "Epi",
  "Epi",
  PDAC_combined$double_positives
)

table(PDAC_combined$double_positives)

Idents(PDAC_combined) <- "double_positives"
PDAC_combined_CAFs_Epi <- subset(PDAC_combined, idents = c("agf", "agf_icaf", "Epi", "icaf", "mycaf", "undefined_caf"))
```

```{r}
SpatialDimPlot(PDAC_combined_CAFs_Epi, 
               images = "slice1",
               group.by = "double_positives", 
               pt.size.factor = 3000,
               cols = c("agf" = 'red', icaf = 'blue', 'mycaf' = 'green', 
                        'undefined_caf' = 'orange','Epi' = 'purple',
                        'agf_icaf' = 'lightblue'),
               alpha = c(1, 1))
```

```{r}
#saveRDS(PDAC_combined_CAFs_Epi, "../Revisions/spatial/subsetted_CAFs_lenient.RDS")


img <- PDAC_combined@images[[1]]
spot_px <- img@scale.factors$spot * ncol(img@image)  # spot diameter in pixels
microns_per_pixel <- 55 / spot_px                          # µm per pixel

cat(sprintf("Spot diameter:      %.1f pixels = 55 µm\n", spot_px))
cat(sprintf("Scale:              %.3f µm per pixel\n",   microns_per_pixel))
cat(sprintf("                    %.3f pixels per µm\n",  1 / microns_per_pixel))

```

##  Spatial feature plots of CAF signature scores

```{r}
# Push AUCell scores from CAFs into PDAC_combined by barcode
auc_cols <- c("mycaf", "icaf", "agf_signature")

for (col in auc_cols) {
  score_vec <- setNames(CAFs@meta.data[[col]], colnames(CAFs))
  PDAC_combined[[col]] <- unname(score_vec[match(colnames(PDAC_combined), names(score_vec))])
}

# sanity check -- should all equal ncol(CAFs)
print(sapply(auc_cols, function(col) sum(!is.na(PDAC_combined[[col]]))))
print(ncol(CAFs))

#now lets look at each tissue site
#now look only at primary tumor 
Idents(PDAC_combined) <- "site"
levels(PDAC_combined)
PT <- subset(PDAC_combined, idents = c("Panc"))
liver <- subset(PDAC_combined, idents = c("Liver"))
peritoneal <- subset(PDAC_combined, idents = c("PC"))
```



### CAF Signature Score vs. Distance from Epi -- ALL 4 signatures, by site

```{r}
library(FNN)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)


# ------------------------------------------------------------------
# Reusable function: distance-to-Epi + AUCell scores for any object
# ------------------------------------------------------------------

get_caf_dist_sig_df <- function(obj) {
  
  coords_list <- lapply(Images(obj), function(img) {
    df <- GetTissueCoordinates(obj, image = img)
    if (nrow(df) == 0) return(NULL)
    df$image <- img
    df
  })
  coords_list <- coords_list[!sapply(coords_list, is.null)]
  if (length(coords_list) == 0) return(NULL)
  coords_full <- do.call(rbind, coords_list)
  
  dist_list <- list()
  
  for (img_name in unique(coords_full$image)) {
    
    cells_in_sample <- rownames(coords_full)[coords_full$image == img_name]
    meta_s <- obj@meta.data[cells_in_sample, , drop = FALSE]
    
    epi_cells <- cells_in_sample[which(meta_s$celltype_cytospace == "Epi")]
    caf_cells <- cells_in_sample[which(!is.na(meta_s$mycaf))]
    
    if (length(epi_cells) < 3 || length(caf_cells) < 3) next
    
    img_obj   <- obj@images[[img_name]]
    spot_px   <- img_obj@scale.factors$spot * ncol(img_obj@image)
    um_per_px <- 55 / spot_px
    
    d <- FNN::get.knnx(
      data  = as.matrix(coords_full[epi_cells, c("imagerow", "imagecol")]),
      query = as.matrix(coords_full[caf_cells, c("imagerow", "imagecol")]),
      k = 1
    )$nn.dist[, 1]
    
    dist_list[[img_name]] <- data.frame(
      barcode  = caf_cells,
      dist_um  = d * um_per_px,
      image    = img_name
    )
  }
  
  caf_dist_df <- bind_rows(dist_list)
  if (nrow(caf_dist_df) == 0) return(NULL)
  
  caf_dist_df$dist_sqrt <- sqrt(caf_dist_df$dist_um)
  
  auc_scores <- obj@meta.data[, auc_cols]
  auc_scores$barcode <- rownames(auc_scores)
  
  caf_dist_df %>% left_join(auc_scores, by = "barcode")
}

# ------------------------------------------------------------------
# Reusable plotting function -- ALL 4 signatures
# ------------------------------------------------------------------

sig_labels <- c(
  "mycaf"         = "myCAF Score",
  "icaf"          = "iCAF Score",
  "agf_signature" = "Neural CAF Score"
)

sig_pal <- c(
  "myCAF Score"      = "#A65628",
  "iCAF Score"       = "#54B0E4",
  "Neural CAF Score" = "#2DAF5A",
  "border"           = "grey50"
)

make_sig_trend_plot <- function(caf_dist_sig_df, title_label) {
  
  d <- caf_dist_sig_df %>%
    filter(dist_sqrt > 0) %>%
    select(dist_sqrt, mycaf, icaf, agf_signature) %>%
    pivot_longer(
      cols      = c("mycaf", "icaf", "agf_signature"),
      names_to  = "Signature",
      values_to = "Score"
    ) %>%
    mutate(Signature = recode(Signature, !!!sig_labels),
           Signature = factor(Signature, levels = c("myCAF Score", "iCAF Score",
                                                 "Neural CAF Score")))
  
  ggplot(d, aes(dist_sqrt, Score, color = Signature)) +
    geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), linewidth = 2) +
    geom_vline(aes(xintercept = 0, color = "border"), linetype = "dashed") +
    xlab("sqrt(Radial Distance from Epi)") +
    ylab("CAF Signature Score") +
    ggtitle(title_label) +
    scale_color_manual(values = sig_pal) +
    theme_bw()
}

# ------------------------------------------------------------------
# Run on your existing PT / liver / peritoneal objects
# ------------------------------------------------------------------

dist_PT         <- get_caf_dist_sig_df(PT)
dist_liver      <- get_caf_dist_sig_df(liver)
dist_peritoneal <- get_caf_dist_sig_df(peritoneal)

site_dist_dfs <- list(
  "Primary Tumor"   = dist_PT,
  "Liver Mets"      = dist_liver,
  "Peritoneal Mets" = dist_peritoneal
)
site_dist_dfs <- site_dist_dfs[!sapply(site_dist_dfs, is.null)]

site_plots <- lapply(names(site_dist_dfs), function(s) {
  make_sig_trend_plot(site_dist_dfs[[s]], title_label = s)
})

wrap_plots(site_plots, nrow = 1) +
  plot_layout(guides = "collect")

ggsave("caf_all4_sig_trends_by_site_scaled_leneint.pdf", width = 6 * length(site_plots), height = 5, dpi = 1200)
dev.off()

#a shared y axis to comapre all three tissue sites
# ------------------------------------------------------------------
# Compute a shared y-axis range across all sites/signatures
# ------------------------------------------------------------------

all_scores <- bind_rows(site_dist_dfs) %>%
  select(mycaf, icaf, agf_signature) %>%
  pivot_longer(everything(), values_to = "Score") %>%
  pull(Score)

y_range <- range(all_scores, na.rm = TRUE)
y_range   # sanity check

# ------------------------------------------------------------------
# Updated plotting function -- takes a shared ylim
# ------------------------------------------------------------------

make_sig_trend_plot <- function(caf_dist_sig_df, title_label, ylim = NULL) {
  
  d <- caf_dist_sig_df %>%
    filter(dist_sqrt > 0) %>%
    select(dist_sqrt, mycaf, icaf, agf_signature) %>%
    pivot_longer(
      cols      = c("mycaf", "icaf", "agf_signature"),
      names_to  = "Signature",
      values_to = "Score"
    ) %>%
    mutate(Signature = recode(Signature, !!!sig_labels),
           Signature = factor(Signature, levels = c("myCAF Score", "iCAF Score",
                                                    "apCAF Score", "Neural CAF Score")))
  
  p <- ggplot(d, aes(dist_sqrt, Score, color = Signature)) +
    geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), linewidth = 2) +
    geom_vline(aes(xintercept = 0, color = "border"), linetype = "dashed") +
    xlab("sqrt(Radial Distance from Epi)") +
    ylab("CAF Signature Score") +
    ggtitle(title_label) +
    scale_color_manual(values = sig_pal) +
    theme_bw() 
  
  if (!is.null(ylim)) {
    p <- p + coord_cartesian(ylim = ylim)
  }
  
  p
}

# ------------------------------------------------------------------
# Rebuild plots with the shared y-axis applied
# ------------------------------------------------------------------

site_plots <- lapply(names(site_dist_dfs), function(s) {
  make_sig_trend_plot(site_dist_dfs[[s]], title_label = s, ylim = c(-1, 2))
})

wrap_plots(site_plots, nrow = 1) +
  plot_layout(guides = "collect")

ggsave("caf_all4_sig_trends_by_site_shared_y_lenient.pdf", width = 6 * length(site_plots), height = 5, dpi = 1200)
dev.off()


##############################################################################
# Spatial feature plots: each CAF signature score over H&E, with Epi (Tumor)
# highlighted in bright green
##############################################################################

library(ggplot2)
library(ggnewscale)

##############################################################################
# Spatial DimPlot: double_positives categories, all image slices
##############################################################################

Idents(PDAC_combined) <- "double_positives"

PDAC_combined_CAFs_Epi <- subset(
  PDAC_combined,
  idents = c("agf", "agf_icaf", "Epi", "icaf", "mycaf", "undefined_caf")
)

double_pos_colors <- c(
  "agf"           = "red",
  "icaf"          = "blue",
  "mycaf"         = "green",
  "undefined_caf" = "orange",
  "Epi"           = "purple",
  "agf_icaf"      = "lightblue"
)
```

```{r}
# ------------------------------------------------------------------
# Single image check
# ------------------------------------------------------------------

SpatialDimPlot(PDAC_combined_CAFs_Epi, 
               images = "slice1.2.13",
               group.by = "double_positives", 
               pt.size.factor = 3000,
               cols = double_pos_colors,
               alpha = c(1, 1))
```

```{r}
# ------------------------------------------------------------------
# All images, one multi-page PDF
# ------------------------------------------------------------------

pdf("double_positives_spatial_all_images.pdf", width = 6, height = 6)
for (img in Images(PDAC_combined_CAFs_Epi)) {
  print(
    SpatialDimPlot(PDAC_combined_CAFs_Epi, 
                   images = img,
                   group.by = "double_positives", 
                   pt.size.factor = 3000,
                   cols = double_pos_colors,
                   alpha = c(1, 1)) +
      ggtitle(img)
  )
}
dev.off()



for (img in Images(PDAC_combined)) {
  cells_in_img <- rownames(GetTissueCoordinates(PDAC_combined, image = img))
  site_here <- unique(PDAC_combined$site[cells_in_img])
  cat(img, "->", paste(site_here, collapse = ", "), "\n")
}
for (img in Images(PDAC_combined)) {
  cells_in_img <- rownames(GetTissueCoordinates(PDAC_combined, image = img))
  site_here <- unique(PDAC_combined$treatment[cells_in_img])
  cat(img, "->", paste(site_here, collapse = ", "), "\n")
}
```